%
% system.tex
%
% Lux Technical Architecture Specification
% Written for principal engineers evaluating the system.
%
% Compile: pdflatex system.tex
%

\documentclass[a4paper,10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!80!black},
  commentstyle=\color{green!50!black},
  stringstyle=\color{red!60!black},
  breaklines=true,
  frame=single,
  framerule=0.3pt,
  rulecolor=\color{gray!50},
  backgroundcolor=\color{gray!5},
  columns=fullflexible,
}

\newcommand{\modpath}[1]{\path{src/punt_lux/#1}}
\newcommand{\tool}[1]{\texttt{#1}}
\newcommand{\msg}[1]{\texttt{#1}}
\newcommand{\elem}[1]{\texttt{#1}}
\newcommand{\field}[1]{\texttt{#1}}

\begin{document}

\title{Lux: Current Architecture Specification}
\author{Visual Output Surface for AI Agents --- a System in Transition}
\date{July 2026 --- v0.19.0}
\hypersetup{
  pdftitle={Lux: Current Architecture Specification},
  pdfauthor={Punt Labs},
  pdfsubject={Architecture},
  pdfcreator={pdflatex}
}
\maketitle

\tableofcontents
\newpage

% ===================================================================
\section{Overview}
% ===================================================================

Lux is a visual output surface for Claude Code agents and local
applications.  Agents and apps push JSON element trees into a native
ImGui window over IPC.  The system has four projection surfaces from a
single codebase: Python library, CLI, MCP server, and Claude Code
plugin.

This document describes the architecture \emph{as the code stands
today}.  Today's code is mid-migration between two models, and this
specification keeps that distinction explicit throughout:

\begin{itemize}
  \item The \textbf{legacy single-process display-server path}
    (\modpath{display/}, \modpath{scene/}, \modpath{socket\_server.py})
    still carries the bulk of the rendering and scene mass.  In this
    path the display process owns the render loop, scene state, widget
    state, interaction queue, and query surface; agents are clients;
    all communication is framed JSON over a Unix domain socket; and the
    loop is single-threaded, polling all I/O non-blocking each frame via
    \texttt{select()}.
  \item The \textbf{new Hub/Display slice} (\modpath{domain/}) is the
    working proof of the target model: a Hub-authoritative
    \texttt{HubDisplay} store, an Element ABC on four migrated kinds, a
    two-tier remote-dispatch event path, and a Hub-scoped
    publish/subscribe channel.  It is narrow but real, and it is where
    new work converges.  See Section \S\ref{sec:hubdisplay}.
\end{itemize}

\subsection{Scope}

This document is the current-architecture record for Lux --- the
org-standard answer to ``what does the code approximately do today?''
It is not the rewrite target.  The canonical design intent lives in
\texttt{docs/architecture/target/target.md}; where this document and
that target disagree, the target defines the \emph{direction} and this
document reports the \emph{present}.

The system is honestly a work in transition.  Rather than pretend the
legacy path is gone or that the new slice is complete, the sections
below label each subsystem as \textbf{migrated}, \textbf{transitional},
or \textbf{legacy} so a reader can tell settled ground from
scaffolding.

\subsection{Companion Documents}

The sibling files in \texttt{docs/architecture/} provide narrower
supporting views:

\begin{itemize}
  \item \texttt{target/target.md} --- the canonical rewrite target.
    Start there for design intent; return here for current behavior.
  \item \texttt{display-server.tex} --- Z specification of the legacy
    display-server state machine (client lifecycle, scene replacement,
    patching, event flush).  It models the single-process path, not the
    Hub/Display runtime, and its refinement tests still hold the current
    display code to that model.
  \item \texttt{workspace-model.tex} --- Z specification of frames,
    client identity, scene-to-frame routing, and World menu ownership.
  \item \texttt{target/introspection-api.md} --- the target
    verification and control surface realized by the unified
    \msg{QueryRequest}/\msg{QueryResponse} envelope and the
    \tool{inspect\_scene()}, \tool{list\_scenes()}, and related tools.
    See Section \S\ref{sec:tools}.
\end{itemize}

\subsection{Design Principles}

\begin{enumerate}[label=\textbf{P\arabic*}.]
  \item \textbf{Composable element tree.}  Every element kind is
    small, nestable, and data-driven.  If an agent can describe it
    as JSON, Lux renders it.
  \item \textbf{Protocol is the API surface.}  The wire protocol is
    the only interface between agents and the display.  No shared
    memory, no function calls, no Python imports required.
  \item \textbf{Single-threaded simplicity.}  The display server
    uses a polling event loop with zero threads.  All state is owned
    by the main thread.  No synchronization primitives.
  \item \textbf{Incremental by default.}  Scenes can be patched by
    element ID without replacing the entire tree.  Table filters run
    at frame rate with zero MCP round-trips.
  \item \textbf{Multi-scene coexistence.}  Each \tool{show()} call
    opens a new tab.  Scenes persist until explicitly dismissed or
    cleared.  Same \field{scene\_id} replaces content in-place.
\end{enumerate}

% ===================================================================
\section{System Architecture}
% ===================================================================

\subsection{Component Map}

The Python package spans four concerns: the protocol layer (the wire
contract), the legacy display/runtime layer with its scene-state
management and MCP/client adapters, and the newer \modpath{domain/}
Hub/Display slice.  The first three are the mass of the shipped
data-display product; \modpath{domain/} is the migrated core described
in Section \S\ref{sec:hubdisplay}.

\begin{longtable}{p{2.3cm}p{3.2cm}p{8.6cm}}
\toprule
\textbf{Module} & \textbf{File or package} & \textbf{Responsibility} \\
\midrule
\endfirsthead
\toprule
\textbf{Module} & \textbf{File or package} & \textbf{Responsibility} \\
\midrule
\endhead
Protocol      & \modpath{protocol/}              & Wire framing, the
                \texttt{FrameReader} streaming buffer, and the
                \texttt{message\_to\_dict} / \texttt{element\_to\_dict}
                dispatch helpers. \\
Elements      & \modpath{protocol/elements/}     & 25 wire element
                kinds plus the \texttt{Patch} payload used by
                \msg{UpdateMessage}.  Four kinds --- \texttt{text},
                \texttt{button}, \texttt{checkbox}, \texttt{dialog} ---
                subclass the \modpath{domain/element\_abc.py}
                \texttt{Element} ABC (data + behavior); the rest are
                frozen dataclasses.  \texttt{JsonElementFactory} decodes
                the io-model kinds, \texttt{ElementCodec} the remainder. \\
Messages      & \modpath{protocol/messages/}     & 22 message
                dataclasses across six families --- \texttt{scene},
                \texttt{lifecycle}, \texttt{menu}, \texttt{introspect},
                \texttt{observer}, and \texttt{remote\_invocation} ---
                with \texttt{MessageRegistry} as the wire dispatch
                table.  User interactions cross the wire as
                \msg{RemoteEventHandlerInvocation}
                (\texttt{remote\_invocation}), not as a broadcast
                interaction message. \\
Display       & \modpath{display/}               & ImGui render loop,
                scene rendering, frame lifecycle, menu management,
                screenshot capture, and event flush.  The core modules
                are \texttt{server.py}, \texttt{element\_renderer.py},
                \texttt{table\_renderer.py}, \texttt{menu\_manager.py},
                \texttt{idle\_screen.py}, and \texttt{texture\_cache.py}. \\
Scene         & \modpath{scene/}                 & Scene graph state.
                \texttt{manager.py} owns \texttt{SceneManager},
                \texttt{frame.py} defines workspace frames, and
                \texttt{widget\_state.py} holds per-scene ImGui control
                state. \\
Domain (Hub)  & \modpath{domain/}                & The migrated
                Hub/Display slice.  \texttt{element\_abc.py} is the
                \texttt{Element} ABC (Composite template method, handler
                registry, property-observer surface).
                \texttt{hub/hub\_display.py} is \texttt{HubDisplay},
                the Hub-authoritative element/owner/child store;
                \texttt{hub/hub.py} is the connection-scoped Agent
                Subscribe pub-sub Hub.  \texttt{display.py} routes
                \texttt{Update}s to \texttt{Event}s and validates
                remote interactions in \texttt{interact()}.
                \texttt{handlers/remote\_dispatch.py} wraps handlers for
                the two-tier path.  See \S\ref{sec:hubdisplay}. \\
Tools (MCP)   & \modpath{tools/}                 & FastMCP tool surface.
                \texttt{server.py} owns the \texttt{FastMCP} instance
                and session lifecycle.  \texttt{tools.py} defines
                display-facing helpers; \texttt{subscribe\_tools.py}
                defines the separate session pub-sub helpers. \\
Socket server & \modpath{socket\_server.py}      & \texttt{SocketServer}
                --- Unix-domain socket lifecycle, non-blocking
                multiplexing of client connections, per-client
                \texttt{FrameReader} buffers, and send/recv to clients. \\
Query dispatcher & \modpath{query\_dispatcher.py} & \texttt{QueryDispatcher}
                --- routes \msg{QueryRequest} by \field{method} string,
                and owns the ring buffers for recent interaction events
                (max 200) and recent error log entries (max 100). \\
Client        & \modpath{client.py}              & Backward-compat shim
                that re-exports \texttt{DisplayClient} under the old
                name \texttt{LuxClient}. \\
Display client & \modpath{display\_client.py}    & \texttt{DisplayClient}
                --- typed \texttt{show}, \texttt{update},
                \texttt{query}, \texttt{clear}, and \texttt{ping}
                methods, plus \texttt{on\_event()} callback registration
                for \msg{RemoteEventHandlerInvocation} frames and
                \texttt{poll\_event()} for queued \msg{ObserverMessage}s. \\
Gateway       & \modpath{luxd.py}                & WebSocket gateway
                (\texttt{luxd}) and mcp-proxy entry point --- the Hub
                process in the target topology.  In the legacy
                single-process path it is transport and session
                bootstrap plumbing, not the source of truth for scene
                state or rendering. \\
Apps/Beads    & \modpath{apps/beads.py}          & Beads issue browser;
                loads \texttt{.beads/issues.jsonl} and renders a
                filterable table scene. \\
Show          & \modpath{show.py}                & \texttt{lux show}
                subcommand for pre-built scenes such as the beads board. \\
Paths         & \modpath{paths.py}               & \texttt{DisplayPaths}
                --- socket discovery, PID lifecycle, and auto-spawn via
                \texttt{ensure()}.  Module-level helpers also resolve
                the optional gateway's PID/port paths. \\
Runtime       & \modpath{runtime.py}             & \texttt{CodeExecutor}
                and \texttt{RenderContext} --- compile and run
                agent-submitted render functions per-frame, gated by
                consent. \\
Config        & \modpath{config.py}              & \texttt{ConfigManager}
                reads and writes YAML frontmatter in
                \texttt{.punt-labs/lux.md}; the live field today is the
                per-repo display-mode toggle (\field{display = y|n}). \\
Hooks         & \modpath{hooks.py}               & Claude Code lifecycle
                hook handlers (session start, post-bash). \\
Service       & \modpath{service.py}             & \texttt{ServiceManager}
                --- launchd/systemd lifecycle for \texttt{luxd}. \\
Remote        & \modpath{remote.py}              & \texttt{ProxyConfigFile}
                --- mcp-proxy WebSocket transport setup. \\
CLI           & \modpath{\_\_main\_\_.py}        & Typer app with
                display, service, status, installation, and gateway
                management subcommands. \\
\bottomrule
\end{longtable}

\subsection{Data Flow}

Three primary data paths connect agents to the display:

\subsubsection{Write Path (Agent $\to$ Display)}

\begin{enumerate}
  \item Agent calls MCP tool (e.g., \tool{show()}).
  \item MCP server obtains a \texttt{DisplayClient} from the session's
    connection registry, then calls
    \texttt{client.show(scene\_id, elements)}.
  \item Client serializes \msg{SceneMessage} as JSON, frames with
    4-byte big-endian length prefix, sends over Unix socket.
  \item \texttt{SocketServer} (\modpath{socket\_server.py})
    receives the frame in \texttt{poll\_clients()}, deserializes via
    \texttt{FrameReader.drain\_typed()}, dispatches to
    \texttt{DisplayServer.\_handle\_message()}.
  \item \texttt{DisplayServer} delegates state updates to
    \texttt{SceneManager.handle\_scene()} or
    \texttt{handle\_framed\_scene()}.
  \item \texttt{DisplayServer.\_render\_scene()} draws the active
    scene each frame via \texttt{ElementRenderer}.
  \item \texttt{DisplayServer} sends an \msg{AckMessage} back to the
    client.
\end{enumerate}

\subsubsection{Event Path (User $\to$ Hub, two-tier D21)}

The event path is the clearest place the migration shows through.  The
display no longer dispatches interactions locally or broadcasts a
generic interaction message; it forwards the interaction back to the
owning Hub, where the authoritative handler runs.  This is the D21
two-tier model described in \S\ref{sec:hubdisplay}.

\begin{enumerate}
  \item When a scene is installed, the display calls
    \texttt{elem.wrap\_handlers\_for\_remote(\_emit\_event)} on each
    element, replacing every handler bucket with a
    \texttt{RemoteDispatchGroup} that captures the socket-send closure.
  \item User interacts with a widget (button click, checkbox toggle).
    \texttt{ElementRenderer} fires the typed event on the display-side
    element copy; the wrapped handler runs instead of the real handler.
  \item The wrapper builds one \msg{RemoteEventHandlerInvocation}
    (\field{element\_id}, \field{action}, \field{event\_kind}
    --- \texttt{button\_clicked} or \texttt{value\_changed} ---,
    \field{value}, \field{ts}) and hands it to \texttt{\_emit\_event},
    which stamps the current \field{scene\_id} and queues it.  One
    interaction produces one message per element/event bucket, however
    many handlers that bucket holds.
  \item Each frame, \texttt{\_flush\_events()} records the events in the
    \texttt{QueryDispatcher} ring buffer (for
    \tool{list\_recent\_events()}), then \texttt{\_dispatch\_queued\_events()}
    routes each event to the \emph{owning} client --- the connection
    whose scene this is, resolved via
    \texttt{SceneManager.scene\_to\_owner} --- falling back to broadcast
    only when no owner is known.
  \item The owning Hub resolves the authoritative Element by
    \field{(scene\_id, element\_id)}, validates ownership in
    \texttt{Display.interact()}, and fires the real event once.  The
    original handler chain --- \texttt{call\_model}, \texttt{publish},
    \texttt{mark\_removed} --- runs on the Hub's copy.
\end{enumerate}

\subsubsection{Update Path (Incremental Patching)}

\begin{enumerate}
  \item Agent calls \tool{update(scene\_id, patches)}.
  \item \texttt{DisplayServer} receives \msg{UpdateMessage}, delegates
    to \texttt{SceneManager.apply\_update()}.
  \item \texttt{\_find\_element()} (module helper in
    \texttt{scene/manager.py}) locates target by ID via depth-first
    search through the element tree.
  \item Patch applied: \field{set} updates fields in-place,
    \field{remove} deletes element and cleans up widget state via
    \texttt{WidgetState.set(eid, None)} and
    \texttt{clear\_suffix()}.
  \item Routes to correct scene by \field{scene\_id} (not limited
    to active tab); \texttt{SceneManager.resolve\_scene()} looks
    up either the unframed-scenes dict or the per-frame scene
    dict.
\end{enumerate}

% ===================================================================
\section{The Hub/Display Slice (Migrated Core)}\label{sec:hubdisplay}
% ===================================================================

Everything above describes the legacy single-process path, still the
mass of the shipped product.  This section describes the migrated core
--- the part of the code that already implements the target model from
\texttt{target/target.md}.  It is narrow but load-bearing: it is where
new element work lands, and it is the reference for how the whole system
is meant to behave once migration completes.

The slice lives in \modpath{domain/} and imports no ImGui, no socket,
and no JSON --- it is pure domain.  Adapters wrap it: the display
process renders it, the Hub transport moves its serialized elements over
the wire.

\subsection{Two Authorities, One Winner}

The target model splits the old display server into a \textbf{Hub} and a
\textbf{Display} that are separate processes, potentially on separate
machines.

\begin{itemize}
  \item The \textbf{Hub} owns authoritative UI state and runs the real
    handlers.  Clients (MCP tools, direct client APIs) talk to the Hub,
    never to the Display.
  \item The \textbf{Display} holds a full replica of the UI for
    rendering and input capture.  It wraps handlers so interactions
    route back to the owning Hub instead of executing locally.
\end{itemize}

The rule is: \textbf{UI state crosses the wire; render calls do not.}
Serialized elements, scene updates, and remote interaction messages
travel between Hub and Display; ImGui calls stay local to the Display.
When Hub-side state changes, the Hub re-sends the whole affected UI and
the Display replaces its copy --- there is no incremental diff protocol.

\subsection{HubDisplay: the Authoritative Store}

\texttt{HubDisplay} (\modpath{domain/hub/hub\_display.py}) is the Hub's
single authoritative surface for scene state.  It is a facade over five
typed collaborators, each with one responsibility:

\begin{center}
\begin{tabular}{lp{8.5cm}}
\toprule
\textbf{Collaborator} & \textbf{Responsibility} \\
\midrule
\texttt{ElementIndex}       & \field{(scene\_id, element\_id)} $\to$
  \texttt{Element} lookup table. \\
\texttt{OwnerTracker}       & Every Element's owning
  \field{ConnectionId}; the interaction path gates on this, and the
  disconnect path walks it to find each connection's owned roots. \\
\texttt{RootRegistry}       & Scene-root ABC Elements that participate
  in the property-observer cascade. \\
\texttt{ChildIndex}         & Parent $\to$ children edges captured at
  install time so cascade removal drops every descendant in one walk. \\
\texttt{HubClientRegistry}  & The set of connections currently
  registered as Hub clients. \\
\bottomrule
\end{tabular}
\end{center}

\texttt{apply(connection\_id, update)} dispatches on the typed
\texttt{Update} sum --- \texttt{AddElement}, \texttt{SetProperty},
\texttt{RemoveElement} --- and enforces ownership: a connection cannot
mutate or evict another connection's elements.  \texttt{AddElement}
recurses into composite children via the Composite Protocol so a Button
buried inside a Dialog is indexed and later click-resolvable.
\texttt{replace\_scene} keeps the Hub authoritative on re-show: it
removes the connection's previous roots, then installs the new ones
through the same \texttt{apply} path so ownership, root observers, and
child indexes rebuild in one place.

\subsection{The Element ABC}

The migrated kinds subclass the \texttt{Element} ABC in
\modpath{domain/element\_abc.py}.  The ABC carries data \emph{and}
behavior:

\begin{itemize}
  \item \textbf{Composite render template.}  \texttt{render()} is a
    template method (never overridden); composites override
    \texttt{\_children()} to return their children, leaves inherit the
    empty default.
  \item \textbf{Handler registry.}  \texttt{add\_handler} /
    \texttt{remove\_handler} / \texttt{fire} key handlers by
    \texttt{Event} subclass.  \texttt{fire} snapshots the bucket so a
    handler that mutates the registry mid-dispatch cannot affect the
    in-flight call.
  \item \textbf{Property-observer surface.}  \texttt{add\_observer},
    \texttt{mark\_removed}, and \texttt{removed} let a parent composite
    react to a child being removed.  All three removal paths --- agent
    \texttt{RemoveElement}, component self-dismiss, connection
    disconnect --- route through the one \texttt{mark\_removed} method.
  \item \textbf{Native serialization.}  \texttt{\_\_reduce\_\_} /
    \texttt{\_\_setstate\_\_} let an Element cross the Hub$\to$Display
    boundary.  Hub-only bookkeeping (the observer closures that
    reference \texttt{HubDisplay}) is dropped on serialization; the
    Display is a replica and does not need Hub observers.  Handlers
    survive so \texttt{wrap\_handlers\_for\_remote} can wrap them on the
    Display side.
\end{itemize}

\subsection{Two-Tier Handler Dispatch (D21)}

The Hub says: ``I am sending you a copy of a UI.  If anyone interacts
with it, I will do the work.''  The Display says: ``I will render that
copy and forward interactions back to the Hub.''

Both tiers hold the same element state and both call
\texttt{elem.fire(event)}.  The difference is what the handler does:

\begin{itemize}
  \item \textbf{Hub:} \texttt{elem.fire(event)} runs the real handler
    --- \texttt{call\_model}, \texttt{publish}, \texttt{mark\_removed}.
  \item \textbf{Display:} \texttt{wrap\_handlers\_for\_remote} has
    replaced each event bucket with a \texttt{RemoteDispatchGroup}
    (\modpath{domain/handlers/remote\_dispatch.py}), so
    \texttt{elem.fire(event)} sends one \msg{RemoteEventHandlerInvocation}
    to the Hub.  The Hub resolves the authoritative Element by
    \field{(scene\_id, element\_id)} in \texttt{Display.interact()},
    validates ownership, and fires the real event once.
\end{itemize}

Two event types cross this boundary today: \texttt{ButtonClicked} and
\texttt{ValueChanged} (\modpath{domain/interaction.py}).  Grouping is
per element/event bucket: a button with three handlers still produces a
single remote message, and the Hub replays the full three-handler chain
once on its copy.  When a handler mutates Hub-side state (for example a
dialog dismissing itself via \texttt{mark\_removed}), the Hub re-pushes
the full scene tree and the Display renders whatever the current tree
is.

\subsection{Agent Subscribe / Publish}

The Hub also owns an application-level publish/subscribe channel
(\modpath{domain/hub/hub.py}), distinct from the element-observer
mechanism above --- the two share no machinery.

Every operation is scoped to a \field{ConnectionId}.  A connection
cannot see, touch, or publish into another connection's topics.
\texttt{subscribe} registers the caller's outbound writer against a
topic; \texttt{publish} fans an \msg{ObserverMessage} payload out to
that connection's subscribers using snapshot-then-iterate so a slow
handler cannot stall concurrent publishes.  Topics --- \texttt{openTicket},
\texttt{closeTicket}, \texttt{markTicketInProgress} --- are defined by
the app or agent, not by Lux.  This is the business-event channel; it is
not the UI-state replication path.

The MCP surface exposes this channel through \tool{subscribe},
\tool{unsubscribe}, \tool{publish}, and \tool{recv} (\S\ref{sec:tools}).

\subsection{Migrated vs.\ Still-Legacy}

To keep the migration boundary honest:

\begin{center}
\begin{tabular}{lp{9.5cm}}
\toprule
\textbf{State} & \textbf{What it covers} \\
\midrule
\textbf{Migrated}     & The Element ABC and its four kinds
  (\texttt{text}, \texttt{button}, \texttt{checkbox}, \texttt{dialog});
  \texttt{HubDisplay} authoritative state with ownership and cascade
  removal; the D21 \msg{RemoteEventHandlerInvocation} two-tier event
  path; Agent Subscribe/Publish. \\
\textbf{Transitional} & The display process, which now runs the D21
  event path (\S\ref{sec:frames}) yet still owns rendering, scene tabs,
  frames, and the query surface as in the legacy design. \\
\textbf{Legacy}       & The 21 not-yet-migrated element kinds as frozen
  dataclasses; \texttt{SceneManager} multi-scene/frame state;
  \texttt{socket\_server.py} single-process transport; the legacy
  ad-hoc introspection request/response pairs kept for compatibility. \\
\bottomrule
\end{tabular}
\end{center}

The direction is fixed by \texttt{target/target.md}; the pace is set by
migrating one element family and one subsystem at a time.

% ===================================================================
\section{Wire Protocol}
% ===================================================================

\subsection{Framing}

Every message is a length-prefixed JSON frame:

\begin{lstlisting}
+------------------+----------------------------+
| 4 bytes (uint32) |  N bytes (UTF-8 JSON)      |
| big-endian       |  payload                   |
+------------------+----------------------------+
\end{lstlisting}

Maximum payload size: 16\,MiB ($2^{24}$ bytes).  The
\texttt{FrameReader} (in \modpath{protocol/\_\_init\_\_.py}) accumulates
partial reads into a byte buffer and yields complete frames via
\texttt{drain()} (raw dicts) or \texttt{drain\_typed()} (Message
dataclasses).  Oversize frames raise \texttt{ValueError}, which
\texttt{SocketServer} converts to a client disconnect.

\subsection{Message Types}

The message layer is split across six families: \texttt{scene},
\texttt{lifecycle}, \texttt{menu}, \texttt{introspect},
\texttt{observer}, and \texttt{remote\_invocation}.  There is no
standalone \texttt{interaction} family --- user interactions travel as
\msg{RemoteEventHandlerInvocation} from the \texttt{remote\_invocation}
module.  In \modpath{protocol/messages/\_\_init\_\_.py}, the protocol
surface is assembled into a \texttt{ClientMessage} union (12
request/command types), a \texttt{DisplayMessage} union (9
response/event types), and \msg{UnknownMessage} for forward
compatibility.

\subsubsection{Client $\to$ Display}

\begin{center}
\begin{tabular}{lp{9cm}}
\toprule
\textbf{Type} & \textbf{Purpose} \\
\midrule
\msg{SceneMessage}  & Replace or add scene.  Fields: \field{id},
  \field{elements[]}, \field{title}, \field{layout},
  \field{frame\_id}, \field{frame\_title}, \field{frame\_size},
  \field{frame\_flags}, \field{frame\_layout}. \\
\msg{UpdateMessage} & Incremental patches.  Fields: \field{scene\_id},
  \field{patches[]} (each: \field{id}, \field{set}, \field{remove}). \\
\msg{ClearMessage}  & Remove all scenes and reset state. \\
\msg{PingMessage}   & Heartbeat.  Field: \field{ts}. \\
\msg{ConnectMessage} & Client identity handshake.  Field: \field{name}. \\
\msg{MenuMessage}   & Set agent-provided menu entries. \\
\msg{ThemeMessage}  & Apply theme by name. \\
\msg{RegisterMenuMessage} & Register per-client menu items.
  Field: \field{items[]}. \\
\msg{IntrospectRequest} & Request scene element tree (legacy
  ad-hoc op; retained alongside \msg{QueryRequest} for
  backward compatibility).
  Field: \field{scene\_id}. \\
\msg{ListScenesRequest} & Request list of all active scenes (legacy
  ad-hoc op). \\
\msg{ScreenshotRequest} & Request display screenshot (legacy
  ad-hoc op). \\
\msg{QueryRequest}  & Generic introspection/control request.  Fields:
  \field{method}, \field{params}.  The dispatcher routes by
  \field{method} string --- see \texttt{introspection-api.md}. \\
\bottomrule
\end{tabular}
\end{center}

\textbf{SceneMessage frame fields} (since v0.9.0): \msg{SceneMessage}
includes optional fields \field{frame\_id}, \field{frame\_title},
\field{frame\_size} (initial \texttt{[width, height]}),
\field{frame\_flags} (dict of ImGui window flags such as
\texttt{no\_resize}, \texttt{no\_collapse}, \texttt{no\_title\_bar},
\texttt{no\_background}, \texttt{no\_scrollbar}, \texttt{auto\_resize}),
and \field{frame\_layout} (\texttt{"tab"} or \texttt{"stack"}).

\subsubsection{Display $\to$ Client}

\begin{center}
\begin{tabular}{lp{9cm}}
\toprule
\textbf{Type} & \textbf{Purpose} \\
\midrule
\msg{ReadyMessage}       & Handshake on connect; protocol version,
  capabilities. \\
\msg{AckMessage}         & Acknowledges scene or update; \field{scene\_id},
  \field{ts}, optional \field{error}. \\
\msg{RemoteEventHandlerInvocation} & User interaction forwarded to the
  owning Hub for authoritative dispatch; \field{element\_id},
  \field{action}, \field{event\_kind}
  (\texttt{button\_clicked} / \texttt{value\_changed}), \field{value},
  \field{ts}, optional \field{scene\_id}. \\
\msg{ObserverMessage}    & Connection-scoped business event fan-out for
  the Agent Subscribe / Publish path; \field{topic} +
  \field{payload}. \\
\msg{PongMessage}        & Ping response; round-trip latency probe. \\
\msg{IntrospectResponse} & Scene element tree response. \\
\msg{ListScenesResponse} & List of active scene summaries plus
  frame summaries (\field{scenes[]}, \field{frames[]}). \\
\msg{ScreenshotResponse} & Path to captured PNG. \\
\msg{QueryResponse}      & Generic introspection/control response.
  Fields: \field{method}, \field{result}, optional \field{error}. \\
\bottomrule
\end{tabular}
\end{center}

Unknown message types are deserialized as \msg{UnknownMessage} for
forward compatibility rather than disconnecting the client.

The dispatch table (\texttt{MessageRegistry} in
\modpath{protocol/messages/registry.py}) is populated at import time by
\texttt{register\_codecs} calls from each family module.  Tests can
construct an isolated registry without touching the module-level
default --- this replaces an earlier \texttt{if/elif} chain in
\texttt{message\_from\_dict}.

\subsection{Element Kinds}

The wire layer exposes 25 element kinds plus the non-element
\texttt{Patch} payload used by \msg{UpdateMessage}.  Four kinds ---
\texttt{text}, \texttt{button}, \texttt{checkbox}, \texttt{dialog} ---
have been migrated onto the \modpath{domain/element\_abc.py}
\texttt{Element} ABC: they own data \emph{and} behavior (a handler
registry, the Composite render template, and the property-observer
surface).  The remaining kinds are still frozen, slotted dataclasses
registered into \texttt{ElementCodec}.  The \texttt{Element} union is
assembled in \modpath{protocol/elements/\_\_init\_\_.py}.

On the encode side, \texttt{\_element\_to\_dict} routes \texttt{text},
\texttt{button}, and \texttt{dialog} through the io-model
\texttt{JsonEncoderFactory}; \texttt{checkbox} is ABC-backed but
serializes through its own per-kind codec.  On the decode side,
\texttt{JsonElementFactory} reconstructs the io-model kinds with the
tier-injected renderer factory and emit callback so handlers wire up on
both Hub and Display.

\begin{longtable}{llp{6cm}}
\toprule
\textbf{Category} & \textbf{Element} & \textbf{Notes} \\
\midrule
\endfirsthead
\toprule
\textbf{Category} & \textbf{Element} & \textbf{Notes} \\
\midrule
\endhead
\texttt{basics}    & \elem{text}              & Content + style (heading, caption, success, error, code);
              optional \field{color} hex string \\
            & \elem{markdown}          & CommonMark rendering via imgui\_md \\
            & \elem{image}             & File path or base64 data \\
            & \elem{separator}         & Visual divider \\
            & \elem{progress}          & Fraction bar with label \\
            & \elem{spinner}           & Loading indicator \\
\midrule
\texttt{inputs}    & \elem{button}            & Label, action, disabled flag;
              \field{arrow} (directional) and \field{small} variants \\
            & \elem{slider}            & Min/max/value, integer mode \\
            & \elem{checkbox}          & Boolean toggle; the shipped
              interactive ABC element --- a toggle fires
              \texttt{ValueChanged}, routed to the Hub as a
              \texttt{value\_changed} invocation \\
            & \elem{combo}             & Dropdown selection \\
            & \elem{input\_text}       & Single-line text input \\
            & \elem{radio}             & Radio button group \\
            & \elem{input\_number}     & Numeric input with step buttons, min/max clamping,
              integer mode \\
            & \elem{color\_picker}     & RGB/RGBA hex picker;
              \field{alpha} (RGBA) and \field{picker} (full widget) modes \\
            & \elem{selectable}        & Click-to-select item \\
\midrule
\texttt{dialog}    & \elem{dialog}            & Dialog root with title,
              children, handlers, and an internal model-backed
              lifecycle \\
\midrule
\texttt{layout}    & \elem{group}             & Row, column, or paged layout
              (\field{pages} + \field{page\_source}) \\
            & \elem{tab\_bar}          & Tabbed container \\
            & \elem{collapsing\_header} & Expandable section \\
            & \elem{window}            & Movable, resizable floating panel \\
            & \elem{tree}              & Recursive node hierarchy;
              \field{flat} flag for non-indented expansion \\
            & \elem{modal}             & Modal popup dialog; container with children,
              emits \texttt{"closed"} event \\
\midrule
\texttt{graphics}  & \elem{draw}              & Canvas with typed draw commands
              (see \S\ref{sec:draw-commands}) \\
            & \elem{plot}              & ImPlot: line, scatter, bar series \\
\midrule
\texttt{table}     & \elem{table}             & Columns, rows, built-in filters, detail panel, pagination \\
\bottomrule
\end{longtable}

\subsection{Draw Commands}\label{sec:draw-commands}

The \elem{draw} element holds a tuple of typed \texttt{DrawCommand}
records.  Each command is a frozen, slotted
dataclass that satisfies the \texttt{DrawCommand} Protocol
(\texttt{TYPE: ClassVar[str]}, \texttt{to\_wire(self) -> dict},
\texttt{from\_wire(cls, d, *, ctx) -> Self}).

\begin{center}
\begin{tabular}{llp{6cm}}
\toprule
\textbf{Module} & \textbf{Record} & \textbf{Geometry} \\
\midrule
\texttt{draw\_commands\_line}   & \texttt{Line}        & Start, end, color, thickness. \\
                                & \texttt{Polyline}    & Point list, color, thickness, closed flag. \\
\midrule
\texttt{draw\_commands\_shape}  & \texttt{Rect}        & Top-left, bottom-right, color, thickness, rounding, filled. \\
                                & \texttt{Circle}      & Center, radius, color, thickness, filled. \\
                                & \texttt{Triangle}    & Three points, color, thickness, filled. \\
\midrule
\texttt{draw\_commands\_curve}  & \texttt{BezierCubic} & Four control points, color, thickness. \\
\midrule
\texttt{draw\_commands\_text}   & \texttt{TextGlyph}   & Position, content, color. \\
\bottomrule
\end{tabular}
\end{center}

Shared value classes live in \texttt{draw\_values.py} and
\texttt{draw\_bounds.py}: \texttt{Point2} (validated $[x, y]$ pair),
\texttt{Color} (\texttt{\#RRGGBB} / \texttt{\#RRGGBBAA} hex),
\texttt{Thickness} (strictly positive stroke width), \texttt{Radius}
(strictly positive), \texttt{Rounding} (non-negative).  The
\texttt{WireContext} in \texttt{draw\_wire.py} carries the per-error
prefix (\texttt{draw command [i] (kind)}) and the require-or-raise
helpers (\texttt{require\_number}, \texttt{require\_sequence}, etc.).
\texttt{DrawCommandDecoder} in \texttt{draw\_decoder.py} dispatches on
the wire \texttt{cmd} string to the appropriate
\texttt{from\_wire} classmethod; \texttt{DrawElement.commands} is
typed \texttt{tuple[DrawCommand, ...]} and the renderer dispatches
via a typed \texttt{match} statement.

% ===================================================================
\section{Display Server Internals}
% ===================================================================

\subsection{Frame Cycle}

The display server runs an ImGui render loop via
\texttt{hello\_imgui.run()}.  Each frame, the
\texttt{DisplayServer.\_on\_frame()} callback executes four phases in
sequence:

\begin{enumerate}
  \item \textbf{Accept connections.}
    \texttt{SocketServer.accept\_connections()} runs \texttt{select()}
    on the server socket with timeout~0.  Accept if readable.  Send
    \msg{ReadyMessage} to the new client.
  \item \textbf{Poll clients.}
    \texttt{SocketServer.poll\_clients()} runs \texttt{select()} on
    all client sockets with timeout~0.  Read available data into the
    per-client \texttt{FrameReader}.  Dispatch complete messages to
    \texttt{DisplayServer.\_handle\_message()}, which routes by
    \texttt{isinstance} to the per-message-type handler.  Remove
    errored sockets via \texttt{remove\_client()}.
  \item \textbf{Render scene.}  Enable docking via
    \texttt{dock\_space\_over\_viewport}.  Render the idle flame
    background (\texttt{idle\_screen.render\_idle()}).  Check for World
    menu background click.  Render World panel if open.  Render all
    workspace frames (see \S\ref{sec:frames}).  Render the dock bar for
    minimized frames.  Legacy fallback: if no frames exist, render
    unframed scenes as tabs.
  \item \textbf{Flush events.}  Route all buffered
    \msg{RemoteEventHandlerInvocation}s to each scene's owning client
    via \texttt{SocketServer.send\_to\_client()} (falling back to
    broadcast only when no owner is known), then clear the queue.  Each
    event is also recorded in
    \texttt{QueryDispatcher.\_recent\_events} for later
    \tool{list\_recent\_events()} queries.
\end{enumerate}

All four phases are non-blocking.  The frame completes in
microseconds when idle.  ImGui handles frame pacing via GLFW vsync;
\texttt{hello\_imgui.fps\_idling.fps\_idle = 30.0} caps the idle
frame rate.

\subsection{Multi-Scene State}

Scene state is owned by \texttt{SceneManager} in
\modpath{scene/manager.py}, exposed to the rendering layer through
read-only properties on \texttt{DisplayServer.scene\_manager}.

\begin{center}
\begin{tabular}{lp{8cm}}
\toprule
\textbf{Field} & \textbf{Type and Purpose} \\
\midrule
\texttt{\_scenes}              & \texttt{dict[str, SceneMessage]} --- unframed
  scenes; ordered by insertion; keyed by \field{scene\_id}. \\
\texttt{\_scene\_order}        & \texttt{list[str]} --- explicit tab ordering for
  unframed scenes. \\
\texttt{\_active\_tab}         & \texttt{str | None} --- currently selected tab in
  the unframed-scenes view. \\
\texttt{\_scene\_widget\_state} & \texttt{dict[str, WidgetState]} --- per-scene
  slider/checkbox/combo/text state. \\
\texttt{\_dirty\_windows}      & \texttt{set[str]} --- window element IDs
  needing initial position set. \\
\midrule
\multicolumn{2}{l}{\textbf{Frame state}} \\
\midrule
\texttt{\_frames}             & \texttt{dict[str, Frame]} --- all workspace
  frames, keyed by \field{frame\_id}. \\
\texttt{\_focus\_frame\_id}   & \texttt{str | None} --- frame to auto-focus
  on next render. \\
\texttt{\_scene\_to\_frame}   & \texttt{dict[str, str]} --- scene
  $\to$ frame lookup. \\
\texttt{\_scene\_to\_owner}   & \texttt{dict[str, int]} --- scene
  $\to$ client fd ownership. \\
\midrule
\multicolumn{2}{l}{\textbf{Display-server state}} \\
\midrule
\texttt{\_event\_queue}       & \texttt{list[RemoteEventHandlerInvocation]}
  on \texttt{DisplayServer}, drained each frame. \\
\texttt{\_client\_names}      & On \texttt{SocketServer}, fd $\to$
  display name from \msg{ConnectMessage}. \\
\texttt{\_menu\_registrations} & On \texttt{MenuManager}, per-client
  agent menu items. \\
\bottomrule
\end{tabular}
\end{center}

\subsubsection{Scene Lifecycle}

\begin{description}[style=nextline]
  \item[New scene (\field{scene\_id} not in \texttt{\_scenes})]
    Appended to \texttt{\_scenes} and \texttt{\_scene\_order}.
    Fresh \texttt{WidgetState} allocated.  Becomes the active tab.
    Window elements marked dirty.  Existing events preserved.

  \item[Replace-in-place (same \field{scene\_id})]
    Content replaced in \texttt{\_scenes}.  Widget state and render
    function state cleared for that scene.  Events for elements
    that existed in the old scene but not the new scene are drained
    via \texttt{on\_scene\_replaced} callback (using full subtree ID
    collection through \texttt{\_collect\_ids()}).  Events from
    other scenes are untouched.

  \item[Dismiss (tab close button)]
    Scene removed from all state dicts.  Window element IDs removed
    from \texttt{\_dirty\_windows}.  Active tab moves to the adjacent
    neighbor (next tab, or previous if rightmost was dismissed).

  \item[Clear (\msg{ClearMessage} or menu)]
    All scenes, widget state, events, and dirty windows cleared.
    Fresh instances allocated (not \texttt{.clear()} on aliased
    objects).
\end{description}

\subsubsection{Widget State Swap Pattern}

The renderer's per-kind \texttt{ElementRenderer.\_RENDERERS} dispatch
table
all reads from \texttt{self.\_widget\_state}.  Rather than threading
a \field{scene\_id} parameter through every method, the display
swaps the per-scene state dict into this instance variable before
rendering each tab:

\begin{lstlisting}[language=Python]
def _render_scene_tab(self, scene_id: str) -> None:
    self._element_renderer.widget_state = \
        self._scene_manager.scene_widget_state[scene_id]
    self._element_renderer.current_scene_id = scene_id
    scene = self._scene_manager.scenes[scene_id]
    for elem in scene.elements:
        self._element_renderer.render_element(elem)
\end{lstlisting}

This is the same pattern ImGui itself uses: ``current context.''
Zero changes required in any renderer method.

\subsection{Table Filtering}

Tables support built-in filtering at frame rate with zero MCP
round-trips.  \texttt{TableRenderer} (in
\modpath{display/table\_renderer.py}) renders filter controls (search
input, combo dropdowns) above the table and applies filters in the
render loop:

\begin{enumerate}
  \item Filter state stored in \texttt{WidgetState} with prefixed
    keys (e.g., \texttt{\_\_tbl\_search\_\{eid\}},\\
    \texttt{\_\_tbl\_filter\_\{col\}\_\{eid\}}).
  \item \texttt{\_apply\_table\_filters()} renders controls, snapshots
    state, detects changes.
  \item \texttt{\_filter\_indexed\_rows()} applies AND logic across
    all active filters (case-insensitive substring for search,
    exact match for combos).
  \item Pagination resets to page~0 when filters change.
  \item Detail panel renders a two-column metadata grid plus
    long-form body text for the selected row.
\end{enumerate}

% ===================================================================
\section{Frame Architecture}\label{sec:frames}
% ===================================================================

Since v0.9.0, scenes can target named \emph{frames}---persistent ImGui
windows that outlive client connections.  Frames are the primary unit
of workspace organization.  The \texttt{Frame} class lives in
\modpath{scene/frame.py}; \texttt{SceneManager} owns the
\texttt{\_frames} dict and the scene-to-frame routing.

\subsection{Frame Class}

\begin{center}
\begin{tabular}{lp{8cm}}
\toprule
\textbf{Field} & \textbf{Type and Purpose} \\
\midrule
\field{frame\_id}       & \texttt{str} --- Unique identifier. \\
\field{title}           & \texttt{str} --- Window title bar text. \\
\field{owner\_fds}      & \texttt{set[int]} --- File descriptors of contributing
  clients. \\
\field{scenes}          & \texttt{dict[str, SceneMessage]} --- All scenes in
  this frame, keyed by \field{scene\_id}. \\
\field{scene\_order}    & \texttt{list[str]} --- Tab ordering. \\
\field{active\_tab}     & \texttt{str | None} --- Currently visible scene. \\
\field{minimized}       & \texttt{bool} --- Docked to bottom bar. \\
\field{cascade\_index}  & \texttt{int} --- Stagger offset for initial
  positioning. \\
\field{initial\_size}   & \texttt{tuple[int,int] | None} --- Width/height hint
  (first use only). \\
\field{flags}           & \texttt{dict[str,bool] | None} --- ImGui window flags
  (\texttt{no\_resize}, \texttt{no\_collapse}, \texttt{no\_title\_bar},
  \texttt{no\_background}, \texttt{no\_scrollbar}, \texttt{auto\_resize}). \\
\field{layout}          & \texttt{Literal["tab", "stack"]} --- Multi-scene
  composition mode. \\
\bottomrule
\end{tabular}
\end{center}

Public read/write properties wrap each field; the constructor follows
the PY-CC-1 \texttt{\_\_new\_\_} pattern.

\subsection{Frame Lifecycle}

\begin{description}[style=nextline]
  \item[Creation]
    A frame is created on the first \msg{SceneMessage} with a
    \field{frame\_id} not already in \texttt{\_frames}.  The client's
    fd is added to \field{owner\_fds}.

  \item[Shared ownership]
    Multiple clients can contribute scenes to the same frame.  Each
    client that sends a scene targeting the frame is added to
    \field{owner\_fds}.

  \item[Orphan model]
    When a client disconnects, its scenes persist with an
    \texttt{\_ORPHAN\_FD} sentinel (\texttt{-1}) in place of the
    real fd.  The frame remains visible and interactive.  A new
    client connecting with the same \field{frame\_id} adopts the
    orphaned scenes.

  \item[Dismissal]
    The close button removes the frame and all its scenes from state.
    Widget state for all scenes in the frame is cleaned up.

  \item[Focus management]
    New frames are positioned using cascade staggering
    (\field{cascade\_index} increments per frame).  Initial size is
    applied on first render only.  \texttt{\_focus\_frame\_id}
    triggers \texttt{SetNextWindowFocus} on the next render cycle.
\end{description}

\subsection{World Menu}

Right-clicking the docking background opens the World menu panel,
providing frame management operations: list all frames, restore
minimized frames, close all frames, and access display settings.

\subsection{Dock Bar}

Minimized frames appear as clickable buttons in a bottom dock bar.
Clicking a dock bar button restores the frame to its previous
position and size.

% ===================================================================
\section{IPC and Process Management}\label{sec:ipc}
% ===================================================================

This section describes the legacy single-process transport: agents and
apps connect to the display over a Unix socket, either directly or
through higher-level client wrappers, and \texttt{luxd} acts as a
WebSocket/session gateway.  In this path scene storage, widget state,
event routing, and rendering all belong to the display process.  The
target topology relocates authoritative scene state into the Hub
(\S\ref{sec:hubdisplay}); the socket mechanics below are what the
migration is moving away from.

\subsection{Socket Discovery}

\texttt{DisplayPaths} in \modpath{paths.py} resolves the display
socket path in priority order:

\begin{enumerate}
  \item \texttt{\$LUX\_SOCKET} environment variable.
  \item \texttt{\$XDG\_RUNTIME\_DIR/lux/display.sock}.
  \item \texttt{/tmp/lux-\$USER/display.sock} (fallback).
\end{enumerate}

The fallback path is chosen to stay within macOS's $\sim$104-character
\texttt{AF\_UNIX} path limit.

The hub socket and PID files are resolved by module-level helpers in
the same file: \texttt{hub\_dir()}, \texttt{hub\_pid\_path()},
\texttt{hub\_port\_path()}, \texttt{hub\_log\_dir()},
\texttt{read\_hub\_port()}, \texttt{is\_hub\_running()}.

\subsection{PID Lifecycle}

\begin{itemize}
  \item PID file written at \texttt{\{socket\_path\}.pid} on server
    startup.
  \item Clients verify liveness via \texttt{os.kill(pid, 0)} before
    connecting.
  \item Stale socket cleanup: \texttt{DisplayPaths.cleanup\_stale()}
    removes socket and PID file if the process is dead.
\end{itemize}

\subsection{Auto-Spawn}

\texttt{DisplayPaths.ensure()} spawns the display server if
not running:

\begin{enumerate}
  \item Checks socket existence and PID liveness.
  \item If not running, spawns\\
    \texttt{python -m punt\_lux display -{}-socket \{path\}}\\
    with \texttt{start\_new\_session=True} (detached from caller's
    process group).
  \item Polls socket existence + PID liveness every 100\,ms until
    ready or 5\,s timeout.
  \item Display logs to \texttt{\{socket\_path\}.log}.
\end{enumerate}

\subsection{Tool Reconnect Path}

The MCP surface wraps display-bound calls in
\texttt{client\_registry.with\_reconnect()}: if an \texttt{OSError}
(for example a broken pipe) occurs, the stale client is closed, a new
connection is established, and the call is retried exactly once.  There
is no unbounded retry loop.

% ===================================================================
\section{Performance}
% ===================================================================

\subsection{Frame Budget}

\begin{center}
\begin{tabular}{lrl}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Notes} \\
\midrule
Idle FPS            & 30             & \texttt{fps\_idling.fps\_idle = 30.0} \\
Active FPS          & 60+            & GLFW vsync, GPU-bound \\
Socket poll timeout & 0              & Non-blocking \texttt{select()} \\
Max message size    & 16\,MiB        & Hard limit in \texttt{FrameReader} \\
Connection timeout  & 5.0\,s         & \texttt{ensure()} and handshake \\
Texture loading     & Lazy           & First render triggers PIL + OpenGL upload \\
Event ring buffer   & 200 entries    & \texttt{QueryDispatcher.\_recent\_events} \\
Error ring buffer   & 100 entries    & \texttt{QueryDispatcher.\_recent\_errors} \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Critical Path Analysis}

The frame-critical path is:

\begin{enumerate}
  \item \texttt{select()} on $N$ client sockets --- $O(N)$, $N < 64$
    in practice.
  \item \texttt{FrameReader.feed()} --- memcpy into buffer, yield
    frames.  $O(B)$ where $B$ is bytes received.
  \item \texttt{DisplayServer.\_render\_scene()} --- ImGui draw list
    calls via \texttt{ElementRenderer.render\_element()}.  $O(E)$
    where $E$ is total element count across visible scene.
  \item \texttt{\_flush\_events()} --- send $K$ events to $N$
    clients.  $O(K \cdot N)$.
\end{enumerate}

Table filtering is $O(R \cdot C)$ where $R$ is row count and $C$
is column count.  This runs every frame but is fast for tables
under $\sim$10{,}000 rows.  No benchmarks exist for larger tables;
pagination (configurable page size) bounds visible row rendering.

\subsection{Idle Screen}

The idle screen renders $\sim$60 draw calls per frame: 48 radial
lines, 3 glow circles, 3 flame shapes (each $\sim$7 bezier
operations).  This is negligible for a GPU-rendered immediate-mode
UI at 30\,FPS idle.

\subsection{Memory}

\begin{itemize}
  \item \textbf{Scene storage.}  Scenes are Python dataclass trees
    held in memory.  No size limit beyond the 16\,MiB message cap.
    Each scene is a separate dict entry; dismissed scenes are
    garbage-collected.
  \item \textbf{Texture cache.}  OpenGL texture IDs held in
    \texttt{TextureCache} until \texttt{\_on\_exit()}.  No eviction
    policy.  Image-heavy scenes accumulate GPU memory.
  \item \textbf{Widget state.}  Per-scene \texttt{WidgetState} is a
    flat dict.  Cleared on scene replacement.  No growth concern.
  \item \textbf{Event queue.}  Flushed every frame.  Bounded by
    interaction rate ($<$100 events/frame in practice).
  \item \textbf{Ring buffers.}  \texttt{deque(maxlen=200)} for events
    and \texttt{deque(maxlen=100)} for errors; constant memory.
\end{itemize}

\subsection{Known Limitations}

\begin{enumerate}
  \item \textbf{No ImGui ID scoping.}  Widget IDs are derived from
    \field{element\_id} alone, not scoped by scene.  ImGui internal
    state (focus, open/closed) can bleed between tabs with colliding
    IDs.
  \item \textbf{No texture eviction.}  Long-running sessions with
    many images accumulate GPU memory.
  \item \textbf{Filter is per-frame.}  Table filtering re-executes
    every frame even when nothing changed.  Not a problem at current
    scale but would benefit from dirty-flag optimization for
    10{,}000+ row tables.
\end{enumerate}

% ===================================================================
\section{Security}
% ===================================================================

\subsection{Threat Model}

The primary trust boundary is the Unix socket.  The display server
accepts connections from any process that can reach the socket file.
In the default configuration (\texttt{/tmp/lux-\$USER/}), the socket
is protected by filesystem permissions (user-only directory, mode
\texttt{0700}).

\begin{center}
\begin{tabular}{lp{5cm}p{5cm}}
\toprule
\textbf{Threat} & \textbf{Description} & \textbf{Mitigation} \\
\midrule
Socket hijacking    & Another process connects to the socket and
  injects scenes or reads events. & Directory permissions (0700) on
  \texttt{/tmp/lux-\$USER/}. \\
Malformed messages  & Client sends invalid JSON, unknown message type,
  or oversize payload. & Unknown types deserialized as
  \msg{UnknownMessage}; oversize payloads rejected by FrameReader. \\
Render function     & Agent-submitted Python code executes arbitrary
  operations. & Consent dialog with AST warnings.  User must
  explicitly click Allow. \\
Resource exhaustion & Scene with millions of elements or very large
  images. & 16\,MiB message cap.  No element count limit (ImGui
  handles gracefully). \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Render Function Security}

Render functions are the only element kind that executes arbitrary
code.  The security model is \textbf{consent-based, not
sandbox-based}:

\begin{enumerate}
  \item \textbf{AST scanning}: a best-effort visitor flags
    suspicious imports (\texttt{os}, \texttt{subprocess},
    \texttt{socket}, \texttt{ctypes}, \texttt{urllib}, etc.) and
    dangerous builtins (\texttt{eval}, \texttt{exec}, \texttt{open},
    \texttt{\_\_import\_\_}).  This is a \emph{UX signal}, not a
    security boundary.

  \item \textbf{Consent dialog}: an ImGui modal displays the full
    source code with line numbers and any AST warnings.  The user
    must click ``Allow'' to proceed.  Dismissing the modal (Esc) is
    treated as ``Deny.''  Consent is per-element: changing the source
    code resets consent.

  \item \textbf{Execution}: \texttt{CodeExecutor} compiles the
    source and calls \texttt{render(ctx)} each frame.  The
    \texttt{RenderContext} provides a persistent state dict, delta
    time, frame counter, canvas dimensions, and a \texttt{send()}
    callback for emitting events.  Execution is \textbf{not sandboxed}:
    the code runs in the display server's process with full access to
    the Python runtime.  Runtime errors are caught and displayed, not
    propagated.
\end{enumerate}

\textbf{Implication}: render functions should be treated as
user-approved code, equivalent to running a script the user
downloaded.  The consent dialog provides informed consent, not
isolation.

\subsection{Protocol Robustness}

\begin{itemize}
  \item \textbf{Malformed JSON}: client disconnected, not crashed.
    \texttt{SocketServer} logs the error via the \texttt{on\_error}
    callback and removes the client.
  \item \textbf{Unknown message type}: deserialized as
    \msg{UnknownMessage} for forward compatibility.
  \item \textbf{Oversize frame}: \texttt{FrameReader} checks
    \texttt{payload\_len} against \texttt{MAX\_MESSAGE\_SIZE} and
    raises; \texttt{SocketServer} converts to disconnect.
  \item \textbf{Partial reads}: \texttt{FrameReader} accumulates
    bytes across frames; no assumption of complete messages per
    \texttt{recv()}.
  \item \textbf{Identity fields protected}:
    \texttt{SceneManager.apply\_update()} silently drops patches
    that attempt to change \field{id} or \field{kind} on an element.
  \item \textbf{Double-remove idempotent}:
    \texttt{SocketServer.remove\_client()} is safe to call twice on
    the same socket.
\end{itemize}

% ===================================================================
\section{Platform Integration}
% ===================================================================

\subsection{macOS}

\begin{itemize}
  \item \textbf{Dock presence.}  The display server runs as a
    normal macOS Dock app per GLFW's default activation policy
    (Regular).  The Dock icon and Cmd-Tab entry are intentionally
    visible to aid operational debugging.
  \item \textbf{Process name.}  \texttt{setproctitle("Lux")} makes
    the process identifiable in \texttt{ps} and \texttt{top}.  Does
    not affect Activity Monitor's binary name column.
  \item \textbf{Font discovery.}  Prefers Arial Unicode or Helvetica;
    merges Apple Symbols for mathematical/Z notation glyphs and STIX
    Two Math for Mathematical Alphanumeric Symbols
    (U+1D400--1D7FF).
  \item \textbf{AF\_UNIX path limit.}  macOS limits Unix socket
    paths to $\sim$104 characters.\\
    \texttt{tempfile.mkdtemp(prefix="lux-")}
    used in tests; production paths are short by design.
\end{itemize}

\subsection{Linux}

\begin{itemize}
  \item \textbf{Font discovery.}  DejaVu Sans or Noto Sans; merges
    Noto Sans Symbols2 and Noto Sans Math for Unicode coverage.
  \item \textbf{XDG compliance.}  Prefers
    \texttt{\$XDG\_RUNTIME\_DIR/lux/} for socket placement.
  \item \textbf{No Dock equivalent.}  Window managers use the X11
    window title (set to ``Lux'').
\end{itemize}

% ===================================================================
\section{MCP Tool Surface}\label{sec:tools}
% ===================================================================

The display-facing MCP surface is registered primarily in
\modpath{tools/tools.py}.  Simple scene and configuration tools are
declared directly with \texttt{@mcp.tool()}; query-style operations
route through the generic \msg{QueryRequest} envelope where possible.
Separate session-pub-sub tools live in \modpath{tools/subscribe\_tools.py}
and are distinct from the display's UI interaction stream.

\begin{longtable}{lp{8cm}}
\toprule
\textbf{Tool} & \textbf{Signature and Purpose} \\
\midrule
\endfirsthead
\toprule
\textbf{Tool} & \textbf{Signature and Purpose} \\
\midrule
\endhead
\multicolumn{2}{l}{\textbf{Scene management}} \\
\midrule
\tool{show}           & \texttt{(scene\_id, elements, title?, layout?,
  frame\_id?, frame\_title?, frame\_size?, frame\_flags?,
  frame\_layout?)} --- Display a scene built from element dicts. \\
\tool{show\_table}    & \texttt{(scene\_id, columns, rows, filters?,
  detail?, flags?, title?, frame\_id?, frame\_title?)} --- Convenience
  wrapper for a filterable table scene. \\
\tool{show\_dashboard} & \texttt{(scene\_id, metrics?, charts?,
  table\_columns?, table\_rows?, title?, frame\_id?, frame\_title?)}
  --- Convenience wrapper for metrics, charts, and a summary table. \\
\tool{update}         & \texttt{(scene\_id, patches)} --- Incremental
  JSON patches targeting elements by ID. \\
\tool{clear}          & \texttt{()} --- Clear all scenes. \\
\midrule
\multicolumn{2}{l}{\textbf{Display control}} \\
\midrule
\tool{ping}           & \texttt{()} --- Latency probe; returns
  \msg{PongMessage}. \\
\tool{set\_menu}      & \texttt{(menus)} --- Set agent-provided menu
  bar entries. \\
\tool{register\_tool} & \texttt{(label, tool\_id, shortcut?, icon?)}
  --- Register an item in the shared Lux Tools menu. \\
\tool{set\_theme}     & \texttt{(theme)} --- Apply display theme.
  Routes through \msg{QueryRequest}. \\
\midrule
\multicolumn{2}{l}{\textbf{Display configuration}} \\
\midrule
\tool{set\_window\_settings} & \texttt{(opacity?, font\_scale?,
  decorated?, fps\_idle?)} --- Configure display window appearance
  and idle frame rate. \\
\tool{set\_frame\_state} & \texttt{(frame\_id, minimized?)}
  --- Control frame state (minimize/restore). \\
\midrule
\multicolumn{2}{l}{\textbf{Display introspection}} \\
\midrule
\tool{inspect\_scene} & \texttt{(scene\_id)} --- Element tree for a
  scene.  The MCP helper routes through \msg{QueryRequest}; the
  legacy \msg{IntrospectRequest} message remains in the wire protocol
  for compatibility. \\
\tool{list\_scenes}   & \texttt{()} --- All active scenes plus
  frame summaries.  The MCP helper also routes through
  \msg{QueryRequest}. \\
\tool{screenshot}     & \texttt{()} --- Intended to capture the
  display as a PNG.  The wire protocol retains a dedicated
  \msg{ScreenshotRequest}/\msg{ScreenshotResponse} path; the current
  MCP helper is transitional and still reaches the display through the
  query surface. \\
\tool{get\_display\_info} & \texttt{()} --- Display dimensions,
  frame count, client count, FPS, uptime, PID. \\
\tool{get\_window\_settings} & \texttt{()} --- Opacity, font scale,
  decorated, idle FPS. \\
\tool{get\_theme}     & \texttt{()} --- Current theme and available
  themes. \\
\tool{list\_clients}  & \texttt{()} --- Connected clients with
  display names and connect times. \\
\tool{list\_menus}    & \texttt{()} --- Registered menu items, per
  client. \\
\tool{list\_recent\_events} & \texttt{(count?)} --- Recent
  interaction events.  Reads
  \texttt{QueryDispatcher.\_recent\_events} (max 200). \\
\tool{list\_errors}   & \texttt{(count?)} --- Recent error log
  entries.  Reads \texttt{QueryDispatcher.\_recent\_errors} (max
  100). \\
\midrule
\multicolumn{2}{l}{\textbf{Per-repo config helpers}} \\
\midrule
\tool{display\_mode}  & \texttt{(repo)} --- Read the repo-local
  display mode from \texttt{<repo>/.punt-labs/lux.md}. \\
\tool{set\_display\_mode} & \texttt{(mode, repo)} --- Write the
  repo-local display mode and eagerly connect when enabling. \\
\midrule
\multicolumn{2}{l}{\textbf{Session pub-sub helpers}} \\
\midrule
\tool{subscribe}      & \texttt{(topic)} --- Subscribe the calling
  session to a topic within its own scope. \\
\tool{unsubscribe}    & \texttt{(topic)} --- Remove the calling
  session's subscription to a topic. \\
\tool{publish}        & \texttt{(topic, payload?)} --- Fan out a
  business event to subscribers in the calling session's scope. \\
\tool{recv}           & \texttt{(timeout?)} --- Wait for the next
  session-scoped business event.  This is \emph{not} the display UI
  interaction stream. \\
\bottomrule
\end{longtable}

The introspection surface is split between the generic
\msg{QueryRequest}/\msg{QueryResponse} envelope and a small number of
older dedicated request/response pairs kept for compatibility.
At the protocol level, \texttt{screenshot} still expects the dedicated
\msg{ScreenshotRequest}/\msg{ScreenshotResponse} path.

Six handlers are built into \texttt{QueryDispatcher} at construction time:
\texttt{inspect\_scene}, \texttt{list\_scenes}, \texttt{list\_clients},
\texttt{list\_menus}, \texttt{list\_recent\_events},
\texttt{list\_errors}.

Seven more handlers are registered by \texttt{DisplayServer} at
construction time because they touch ImGui state:
\texttt{screenshot} (stub --- raises),
\texttt{get\_display\_info},
\texttt{get\_window\_settings}, \texttt{get\_theme},
\texttt{set\_window\_settings}, \texttt{set\_frame\_state},
\texttt{set\_theme}.

% ===================================================================
\section{Formal Model}\label{sec:formal}
% ===================================================================

The legacy display server's state machine is formally specified in
\texttt{docs/architecture/display-server.tex} using the Z notation
(ISO 13568).  The specification models the single-process path, and its
refinement tests still hold the current display code to that model.  It
covers:

\begin{itemize}
  \item Client connection/disconnection with FrameReader association.
  \item Scene replacement, incremental patching, and clearing.
  \item Interaction event generation and broadcast.
  \item FrameReader streaming buffer invariants.
\end{itemize}

The specification has been type-checked with Spivey's \texttt{fuzz}
and model-checked with ProB (7{,}943 reachable states explored).
Refinement tests in \texttt{tests/test\_display\_refinement.py}
verify commutativity between the abstract Z operations and the
concrete Python implementation via an abstraction function
(\texttt{tests/display\_abstraction.py}).

Partition tests in \texttt{tests/test\_display\_partition.py} derive
test cases from the Z specification using the Test Template Framework
(TTF), covering all input space partitions of each operation schema.

\textbf{Note}: the formal model predates multi-scene tabs and models
a single-scene display.  The abstraction function maps multi-scene
state to the active tab for backward compatibility.  A specification
update to model the full multi-scene state is pending.

% ===================================================================
\section{Test Architecture}
% ===================================================================

The test suite is broad and layered; the exact count changes regularly,
so this document records the coverage shape rather than a brittle
inventory snapshot.

\begin{itemize}
  \item \textbf{Protocol and codec tests}: framing, serialization,
    typed draw-command decode, value validation, and query envelopes.
    Representative files:
    \texttt{test\_protocol.py}, \texttt{test\_draw\_*},
    \texttt{test\_query.py}.
  \item \textbf{Display-state and manager tests}:
    pure state transitions for the display server without a live GUI.
    Representative files: \texttt{test\_display\_state.py},
    \texttt{test\_scene\_manager.py},
    \texttt{test\_query\_dispatcher.py},
    \texttt{test\_socket\_server.py}.
  \item \textbf{Tooling and client tests}: \texttt{test\_tools.py},
    \texttt{test\_display\_client.py}, \texttt{test\_show.py},
    \texttt{test\_cli.py}, \texttt{test\_paths.py},
    \texttt{test\_service.py}, \texttt{test\_remote.py}, and
    \texttt{test\_config.py} cover the MCP surface, client wrappers,
    CLI, process helpers, remote config, and repo-local configuration.
  \item \textbf{Formal-model alignment}:
    \texttt{test\_display\_refinement.py} and
    \texttt{test\_display\_partition.py} tie the concrete Python
    implementation back to the Z specification.
  \item \textbf{Marker-gated integration and e2e tests}:
    \texttt{test\_integration.py}, the integration-marked cases in
    \texttt{test\_query.py}, and \texttt{test\_e2e.py} exercise real
    sockets and, for the e2e path, a live display environment.
\end{itemize}

Quality gates are driven by the Makefile: \texttt{ruff} (lint +
format), \texttt{mypy} (strict), \texttt{pyright} (strict),
\texttt{pytest} (default marker set), \texttt{check-oo} (OO ratchet
against \texttt{.oo-baseline.json}), and
\texttt{check-suppressions}.  \texttt{make check} is the composite
gate.

\end{document}
