%
% hub_replicator.tex
%
% A Z specification of the lux Hub replicator: one background worker that
% is the sole writer to the display connection, several mutator threads
% (MCP tool calls and Hub-side click handlers) that only ever touch the
% Hub's store, a read/write lock on the store, the client send lock, and a
% kill-and-restart path that runs while mutators keep changing the store.
%
% Models the DESIGN in docs/architecture/mcp-display-liveness.md, sections
% "The Hub replicator: one background worker", "The store needs a lock",
% and "Killing a stuck display and starting a new one".  It does NOT model
% the current code, which still carries the store race this design closes.
%
% Type-check:  fuzz docs/hub_replicator.tex
% Model-check (see the Verification section for the exact goal predicates):
%   probcli docs/hub_replicator.tex -model_check -p DEFAULT_SETSIZE 2
%   probcli docs/hub_replicator.tex -model_check -nodead \
%       -goal "readLock = held & clientLock = held" -p DEFAULT_SETSIZE 2
%
% Formatting follows the fuzz house rules: \quad~ for continuation indent,
% schema lines under 80 columns, predicates broken at \land/\lor/\implies.
%

\documentclass[a4paper,10pt,fleqn]{article}
\usepackage[margin=1in]{geometry}
\usepackage{fuzz}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\begin{document}

\title{The lux Hub Replicator: a Z Specification}
\author{Formal model of the single-writer, two-lock replication discipline}
\date{July 2026}
\maketitle

% ============================================================
\section{Introduction}
% ============================================================

An MCP \texttt{clear} call once froze an agent for about 38 minutes: the
tool sent to the display synchronously, on the agent's own thread, over a
socket with no send timeout, while holding the client send lock.  The fix
moves every send to the display onto one background worker inside the Hub
--- the \emph{replicator} --- and keeps every agent-facing tool on the
Hub's in-memory store, which it updates and returns from at once.

The worker introduces real concurrency.  It is a background thread; it
drains a set of changed scenes that several other threads mark; it reads a
store that those threads mutate; and it kills and restarts a wedged
display while those threads keep running.  This document models that
concurrency as a finite state machine over a small fixed set of scenes and
mutator threads, so that ProB can enumerate every interleaving and either
confirm the invariants or produce the exact offending trace.

Six properties must hold across every interleaving:

\begin{description}
  \item[I1 --- single writer.] Only the worker ever writes to the display
    connection.  No mutator thread holds the client send lock.
  \item[I2 --- agent-path liveness.] No mutator operation is ever disabled
    by display I/O or by the client send lock.  A mutator makes progress
    even while the worker is stuck sending to a dead display.
  \item[I3 --- no lock cycle.] No thread ever holds the store lock and the
    client send lock at the same time.  The worker snapshots under the
    store read lock, releases it, and only then takes the client lock to
    send.
  \item[I4 --- no lost update.] Whenever the system comes to rest with the
    display up, the display shows the store's latest state for every scene
    --- including after a reap and respawn, which re-marks every live
    scene, and including after a cycle that failed and restored its own
    batch.
  \item[I5 --- one display.] The worker never spawns a fresh display while
    one is still alive; a respawn follows a kill.
  \item[I6 --- menu-delivery durability.] A menu change that has been
    marked is either delivered to the display by the worker's own send or
    stays marked; a failed cycle never drops it.  When the system comes to
    rest, no marked menu change is left undelivered.
\end{description}

\paragraph{Recovery and blanking.}
The worker's cycle can fail in three ways, and an emptied scene must
disappear from the display; both shape the model.  A send raises one of two
socket errors --- a timeout (\texttt{BlockingIOError}) or a dead peer
(\texttt{OSError}) --- and each has its own heal.  But the heal itself can
fail: an unspawnable display or a refused reconnect raises before the work is
re-marked, and any other error inside the cycle escapes the two socket
handlers.  That third, generic failure must not drop the drained work: the
worker re-marks it --- every scene the cycle drained \emph{and} the clear it
carried \emph{and} the menu flag it carried --- and backs off.  The two socket
heals re-mark more: the whole drained batch, together with every scene the
store still holds live, together with the carried clear and the carried menu
flag.  The drained scenes matter because a batch can drain a scene whose only
job is to be blanked, and a reconnect can land on the same still-running
display that keeps its stale copy, so a lost blank must be re-queued or the
departed scene lingers forever; blank-then-repaint is idempotent.  Finally, a
drained scene whose store copy is empty is pushed as a blank into its own
recorded frame, so an emptied or departed scene vanishes --- unless the batch
already carried a clear, which blanked the whole display, in which case the
empty scene is skipped rather than blanked twice.

\paragraph{The menu leg.}
Alongside the scenes and the clear, the worker also pushes the agent menu
bar and the registered World-menu items.  A menu change is a payload-less
flag, like the clear: a mutator sets it, and the worker reads the menu
registry \emph{fresh} at send time, pushing whatever the registry holds
then.  Reading fresh is the whole point of the payload-less design --- a
menu change that lands while an earlier send is failing is never overwritten
by a stale re-mark, because there is no payload to go stale; the newest
registry state always wins.  The menu push is one send with the same time
limit and the same two socket failures as a scene send, so a wedged or dead
display on the menu push heals exactly as it does on a scene push, and the
carried menu flag is re-marked so the healed display is repainted.

% ============================================================
\section{Basic Types}
% ============================================================

\begin{zed}
  [SCENE, MUTATOR]
\end{zed}

\texttt{SCENE} is the set of scenes the Hub can hold; two suffice to
exhibit a coalesced batch and a per-scene lost update.  \texttt{MUTATOR}
is the set of threads that change the store: MCP tool calls (\texttt{show},
\texttt{update}, \texttt{clear}) and Hub-side click handlers.  Both kinds
do the same thing to the store, so one carrier covers both; two mutators
exhibit lock contention.  Both carriers are bounded by ProB's
\texttt{DEFAULT\_SETSIZE}.

% ============================================================
\section{Free Types}
% ============================================================

\begin{zed}
  VAL ::= vempty | vlive | vtorn
\end{zed}

The content of one scene in the store or on the display, abstracted to a
token.  \texttt{vempty}: no elements (a scene that has been cleared or
never populated).  \texttt{vlive}: a committed, consistent scene tree.
\texttt{vtorn}: the half-updated state a scene passes through while a
mutator is inside a multi-step \texttt{replace\_scene} --- the value that,
if read without the lock, is the ``dictionary changed size during
iteration'' defect.  Real scene structure is elided; what matters is
whether a read observes a committed value or a torn one.

\begin{zed}
  DPROC ::= dup | dstuck | ddead
\end{zed}

The display process as the worker's send sees it.  \texttt{dup}: alive and
draining its socket; a send succeeds.  \texttt{dstuck}: alive but no longer
reading, so the send buffer fills and the time-limited send raises
\texttt{BlockingIOError}.  \texttt{ddead}: the peer is gone, so the send
raises a plain \texttt{OSError} (\texttt{ECONNRESET}/\texttt{EPIPE}).

\begin{zed}
  LOCK ::= held | free
\end{zed}

\begin{zed}
  FLAG ::= set | clear
\end{zed}

Two-valued markers for the worker's read lock and client lock, and for the
cleared, shutting, and menu signals, respectively.  A dedicated free type is
used rather than a built-in boolean so the spec is self-contained for
\texttt{fuzz}.

\begin{zed}
  WPHASE ::= \\
  \quad~ widle | wdrain | wblank | wmenu | wproc | wsnap | wsend | wmsend | \\
  \quad~ wreap | wensure | wremark | wreconn | \\
  \quad~ wflush | wstopped
\end{zed}

The worker's own program counter.  \texttt{widle}: waiting on the
condition.  \texttt{wdrain}: woken, about to take the whole changed-scene
set.  \texttt{wblank}: batch drained, about to blank the display if the
batch carried a clear.  \texttt{wmenu}: display blanked (or not), about to
push the menu registry if the batch carried a menu change.  \texttt{wproc}:
iterating the drained batch.  \texttt{wsnap}: holding the store read lock,
copying one scene's roots.  \texttt{wsend}: holding the client lock, sending
that copy.  \texttt{wmsend}: holding the client lock, sending the menu push.
\texttt{wreap}/\texttt{wensure}/\texttt{wremark}: the reap-respawn-remark
recovery after a send timeout.  \texttt{wreconn}: the reconnect recovery
after a dead peer.  \texttt{wflush}/\texttt{wstopped}: the final bounded
flush at shutdown.  A generic cycle failure returns the worker straight to
\texttt{widle} after restoring its batch.

% ============================================================
\section{State}
% ============================================================

The state is flat: one schema.  Its predicate carries only
\emph{structural} coherence --- the locks and the snapshot hold at most one
thing, and the write lock is held exactly when a scene is mid-replace.
These hold in every design, correct or buggy, so they constrain the state
space without masking the safety properties.  The safety properties
(Section~\ref{sec:inv}) are deliberately \emph{absent} here, so that a
violating state stays reachable and ProB can find it.

\begin{schema}{Repl}
  store : SCENE \fun VAL \\
  displayed : SCENE \fun VAL \\
  dirty : \power SCENE \\
  batch : \power SCENE \\
  drained : \power SCENE \\
  cleared : FLAG \\
  batchCleared : FLAG \\
  menusDirty : FLAG \\
  batchMenus : FLAG \\
  menuShown : FLAG \\
  snapScene : \power SCENE \\
  snapVal : VAL \\
  storeLockMut : \power MUTATOR \\
  mutScene : \power SCENE \\
  readLock : LOCK \\
  clientLock : LOCK \\
  display : DPROC \\
  shutting : FLAG \\
  wphase : WPHASE
\where
  \# storeLockMut \leq 1 \\
  \# snapScene \leq 1 \\
  \# mutScene \leq 1 \\
  (storeLockMut = \emptyset \iff mutScene = \emptyset)
\end{schema}

\texttt{store} is the Hub's authoritative copy; \texttt{displayed} is what
the display is currently showing.  \texttt{dirty} is the changed-scene set
guarded by the worker's condition, and \texttt{cleared} the cleared flag
beside it --- both live signal state that a mutator sets and the worker
resets when it drains.  \texttt{batch} is the set of scenes still to send this cycle --- it shrinks
as each is pushed --- while \texttt{drained} is the whole set the cycle
drained, kept unshrunk so a generic failure can restore all of it.  The two
mirror the shipped \texttt{DrainedBatch}: its \texttt{scenes} field is
immutable (\texttt{drained}), and the worker's loop index over it is what
\texttt{batch} tracks.  The distinction is load-bearing: a generic failure of
a batch that carried a clear must re-mark \emph{every} scene it drained, not
just the unsent remainder, because the re-marked clear will blank the whole
display and every drained scene must then be repainted.  \texttt{batchCleared}
is the clear that \emph{that} drained batch carried: the worker snapshots the
cleared flag into \texttt{batchCleared} when it drains, resets the live flag,
and keeps \texttt{batchCleared} for the whole cycle, so both the empty-scene
skip and a recovery re-mark can still see that the batch was a clearing one.
Modelling the two separately is what makes the recovery re-mark of a
consumed clear expressible; a single flag, reset when the blank is pushed,
would have forgotten by the time a send failed.

\texttt{menusDirty} is the live menu-change flag beside \texttt{cleared}: a
mutator sets it when the menu registry changes, and the worker resets it when
it drains.  \texttt{batchMenus} is the menu flag \emph{that} drained batch
carried --- snapshotted from \texttt{menusDirty} at the drain and kept for the
whole cycle, exactly as \texttt{batchCleared} carries the clear --- so a
recovery re-mark can see that the failed batch owed a menu push.
\texttt{menuShown} records whether the worker has delivered the current menu
registry to the display since the last menu change: \texttt{set} once a menu
push has succeeded, \texttt{clear} once a change is marked and not yet
delivered.  It is a statement about the worker's \emph{delivery obligation},
not about a menu payload, because the menu is payload-less and read fresh at
send time; it is the variable I6 is stated over.

\texttt{snapScene}/%
\texttt{snapVal} are the one scene the worker has copied out under the read
lock and the value it copied.  \texttt{storeLockMut} names the mutator
holding the store write lock (empty or one), and \texttt{mutScene} the scene
it is mid-replacing.  \texttt{readLock} and \texttt{clientLock} record
whether the worker holds the store read lock and the client send lock.
\texttt{display} is the display process state; \texttt{shutting} the
clean-shutdown signal; \texttt{wphase} the worker's program counter.

Modelling the store read/write lock as \emph{one} mutually-exclusive slot
(\texttt{readLock} for the sole reader, \texttt{storeLockMut} for the one
writer) is sound because there is exactly one reader thread in this design
--- the worker.  A read/write lock exists to allow concurrent readers; with
a single reader the only exclusion that can ever fire is reader-versus-%
writer, which a single slot captures.

\texttt{menuShown} is modelled at the level of the worker's delivery, not at
the level of the display process's live menu content, and the two differ in
one honest way.  For scenes, the model resets \texttt{displayed} to empty on a
respawn (\texttt{Ensure}) and proves the content property I4 --- so it must
track what the fresh display shows.  For menus it proves the weaker delivery
property I6 --- ``a marked change is eventually pushed or stays marked'' ---
so it tracks whether a push is owed (\texttt{menusDirty}) and whether the last
owed push landed (\texttt{menuShown}).  The re-establishment of a menu on a
freshly respawned display over the client handshake is a separate subsystem,
outside this replicator's dirty-signal discipline, exactly as the client-side
replay of scenes is outside it; the model therefore leaves \texttt{menuShown}
unchanged across \texttt{Ensure} and \texttt{Reconn} and holds the worker to
its own obligation: never drop a marked menu change.

% ============================================================
\section{Initialization}
% ============================================================

\begin{schema}{Init}
  Repl'
\where
  store' = (\lambda s : SCENE @ vempty) \\
  displayed' = (\lambda s : SCENE @ vempty) \\
  dirty' = \emptyset \\
  batch' = \emptyset \\
  drained' = \emptyset \\
  cleared' = clear \\
  batchCleared' = clear \\
  menusDirty' = clear \\
  batchMenus' = clear \\
  menuShown' = set \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  storeLockMut' = \emptyset \\
  mutScene' = \emptyset \\
  readLock' = free \\
  clientLock' = free \\
  display' = dup \\
  shutting' = clear \\
  wphase' = widle
\end{schema}

A single initial state: empty store, empty display, nothing dirty, no menu
change owed and the menu delivered, no locks held, the worker idle, the
display up.

% ============================================================
\section{Mutator Operations}
% ============================================================

A mutator changes the store under the write lock in two steps, so the
scene passes through a torn intermediate the worker must never read.  The
first step takes the lock and begins the replace; the second commits the
new value, marks the scene dirty, and releases the lock.  Marking dirty is
part of the commit: the tool calls \texttt{mark\_dirty} on the same thread,
immediately after \texttt{replace\_scene} returns and before it yields, so
no other thread observes the store between the store write and the mark.

The write lock is taken only when it is free \emph{and} the worker is not
mid-snapshot (\texttt{readLock = free}).  That guard is the store lock's
reader-writer exclusion; removing it from the worker's snapshot yields the
buggy variant of Section~\ref{sec:fidelity}.

\begin{schema}{ReplaceBegin}
  \Delta Repl \\
  m? : MUTATOR \\
  s? : SCENE
\where
  storeLockMut = \emptyset \\
  readLock = free \\
  storeLockMut' = \{ m? \} \\
  mutScene' = \{ s? \} \\
  store' = store \oplus \{ s? \mapsto vtorn \} \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

\begin{schema}{ReplaceCommit}
  \Delta Repl \\
  m? : MUTATOR
\where
  m? \in storeLockMut \\
  store' = store \oplus (mutScene \cross \{ vlive \}) \\
  dirty' = dirty \cup mutScene \\
  storeLockMut' = \emptyset \\
  mutScene' = \emptyset \\
  displayed' = displayed \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

A single scene can also be emptied without a clear --- an \texttt{update}
that strips a scene to no roots, or a session whose disconnect blanks the
scenes it alone held.  \texttt{EmptyScene} sets one scene's store value to
\texttt{vempty} and marks it dirty, under the same write-lock discipline as
a replace.  It is what makes a departed scene's blank observable to the
worker.

\begin{schema}{EmptyScene}
  \Delta Repl \\
  s? : SCENE
\where
  storeLockMut = \emptyset \\
  readLock = free \\
  store' = store \oplus \{ s? \mapsto vempty \} \\
  dirty' = dirty \cup \{ s? \} \\
  displayed' = displayed \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

The \texttt{clear} tool empties the whole store under the write lock and
sets the cleared flag instead of marking a scene dirty.  Modelled as one
step, since the torn-read race is already exhibited by \texttt{replace}:

\begin{schema}{ClearStore}
  \Delta Repl
\where
  storeLockMut = \emptyset \\
  readLock = free \\
  store' = (\lambda s : SCENE @ vempty) \\
  cleared' = set \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

The \texttt{set\_menu} / \texttt{set\_registered\_items} tools change the
Hub's menu registry and flag the change, exactly as \texttt{clear} sets the
cleared flag.  \texttt{MarkMenus} sets the live menu flag and records that the
display is now behind the registry (\texttt{menuShown' = clear}).  Like the
real \texttt{mark\_menus}, it is queue-only: it touches no store, takes no
lock, and is guarded only by \texttt{menusDirty = clear} --- coalescing, since
a second change while one is already pending is a no-op.  It never mentions
\texttt{clientLock}, \texttt{display}, or \texttt{wphase}, so no state of the
display can disable it; the menu-set path is as free of display I/O as the
store mutators are (I2).

\begin{schema}{MarkMenus}
  \Delta Repl
\where
  menusDirty = clear \\
  menusDirty' = set \\
  menuShown' = clear \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  batchMenus' = batchMenus \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

% ============================================================
\section{The Worker Loop}
% ============================================================

The worker waits on the condition, wakes when something is dirty, the
screen was cleared, or the menu registry changed, drains the whole set, and
pushes each scene by copying its current roots under the read lock and
sending the copy outside the lock.  Coalescing needs no extra mechanism:
many marks of one scene add it to the set once, and the worker reads that
scene's current store value when it snapshots, so the display receives the
newest value, never a half-finished one.

\begin{schema}{Wake}
  \Delta Repl
\where
  wphase = widle \\
  shutting = clear \\
  (dirty \neq \emptyset \lor cleared = set \lor menusDirty = set) \\
  wphase' = wdrain \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\texttt{Drain} takes the whole changed-scene set at once, snapshots the
cleared flag into \texttt{batchCleared} and the menu flag into
\texttt{batchMenus} for the duration of the cycle, and resets both live flags
--- exactly as \texttt{wait\_and\_drain} builds an immutable
\texttt{DrainedBatch} and clears the signal.  A clear or a menu change that
arrives after this point re-sets its live flag for the next cycle.

\begin{schema}{Drain}
  \Delta Repl
\where
  wphase = wdrain \\
  batch' = dirty \\
  drained' = dirty \\
  dirty' = \emptyset \\
  batchCleared' = cleared \\
  cleared' = clear \\
  batchMenus' = menusDirty \\
  menusDirty' = clear \\
  menuShown' = menuShown \\
  wphase' = wblank \\
  store' = store \\
  displayed' = displayed \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

At \texttt{wblank} the worker blanks the display first if the batch carried
a clear, and only then turns to the menu and the scenes.  \texttt{ApplyClear}
pushes the blank; \texttt{SkipBlank} is the no-clear case.  Both advance to
\texttt{wmenu}.  \texttt{batchCleared} is \emph{not} consumed here: it survives
the blank so that the per-scene empty skip and any recovery re-mark can still
see that this was a clearing batch.

\begin{schema}{ApplyClear}
  \Delta Repl
\where
  wphase = wblank \\
  batchCleared = set \\
  displayed' = (\lambda s : SCENE @ vempty) \\
  wphase' = wmenu \\
  store' = store \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\begin{schema}{SkipBlank}
  \Delta Repl
\where
  wphase = wblank \\
  batchCleared = clear \\
  wphase' = wmenu \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

At \texttt{wmenu} the worker pushes the menu registry if the batch carried a
menu change, and otherwise skips straight to the scenes.  The push is one
send, so it takes the client lock and goes to \texttt{wmsend}; the store read
lock is not held, so I3 is trivially preserved on this leg.  Reading the
registry is instantaneous and does not appear in the state, because the menu
is payload-less: whatever the registry holds when \texttt{MenuSendOk} fires is
what the display receives, which is why a stale re-mark can never lose a
newer change.

\begin{schema}{SendMenusBegin}
  \Delta Repl
\where
  wphase = wmenu \\
  batchMenus = set \\
  clientLock' = held \\
  wphase' = wmsend \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\begin{schema}{MenuSkip}
  \Delta Repl
\where
  wphase = wmenu \\
  batchMenus = clear \\
  wphase' = wproc \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

The menu send has the same three outcomes as a scene send, keyed on the
display process.  On success the display now shows the current registry, so
the delivery obligation is met (\texttt{menuShown' = set}) and the worker
turns to the scenes.  A wedged display sends the worker to the reap path; a
dead peer to the reconnect path.  In both failures \texttt{menuShown} stays
\texttt{clear}, so the obligation is still owed until a recovery re-marks it
and a later cycle re-delivers.

\begin{schema}{MenuSendOk}
  \Delta Repl
\where
  wphase = wmsend \\
  display = dup \\
  clientLock' = free \\
  menuShown' = set \\
  wphase' = wproc \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\begin{schema}{MenuSendStuck}
  \Delta Repl
\where
  wphase = wmsend \\
  display = dstuck \\
  clientLock' = free \\
  wphase' = wreap \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\begin{schema}{MenuSendDead}
  \Delta Repl
\where
  wphase = wmsend \\
  display = ddead \\
  clientLock' = free \\
  wphase' = wreconn \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\texttt{Snap} takes the store read lock, picks one scene still in the
batch, and copies its current value.  The read lock is available only when
no mutator holds the write lock (\texttt{storeLockMut = $\emptyset$}); that
guard is what stops the worker reading a torn scene.  Blank-before-repaint
is now enforced by the \texttt{wblank} phase, which precedes \texttt{wproc},
so \texttt{Snap} carries no cleared guard of its own.

\begin{schema}{Snap}
  \Delta Repl
\where
  wphase = wproc \\
  batch \neq \emptyset \\
  storeLockMut = \emptyset \\
  (\exists s : SCENE @ \\
  \quad~ s \in batch \land snapScene' = \{ s \} \land snapVal' = store~s) \\
  readLock' = held \\
  wphase' = wsnap \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

From \texttt{wsnap} the worker either sends the copy or, when the copy is
empty and the batch already blanked the whole display with a clear, skips
it.  \texttt{SendBegin} releases the store read lock and only then takes the
client send lock; this release-before-acquire is the whole of I3.  Its
guard \texttt{$\lnot$(snapVal = vempty $\land$ batchCleared = set)} is the
send/skip partition: an empty scene under a clear is the one case that does
not send.

\begin{schema}{SendBegin}
  \Delta Repl
\where
  wphase = wsnap \\
  \lnot\,(snapVal = vempty \land batchCleared = set) \\
  readLock' = free \\
  clientLock' = held \\
  wphase' = wsend \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  display' = display \\
  shutting' = shutting
\end{schema}

\texttt{SendSkip} is the complementary case: the scene is empty and the
clear already blanked the whole display, so the worker copied nothing worth
sending.  It releases the read lock, drops the scene from the batch, and
sends nothing --- \texttt{displayed} is untouched, since the clear's blank
already set that scene to \texttt{vempty}.

\begin{schema}{SendSkip}
  \Delta Repl
\where
  wphase = wsnap \\
  snapVal = vempty \land batchCleared = set \\
  readLock' = free \\
  batch' = batch \setminus snapScene \\
  drained' = drained \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  wphase' = wproc \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

The time-limited send has three outcomes, keyed on the display process.
On success the display now shows the copied value and the scene leaves the
batch.  When the copied value is \texttt{vempty} --- an emptied or departed
scene sent without a clear --- this is the blank that makes it vanish from
its own frame.

\begin{schema}{SendOk}
  \Delta Repl
\where
  wphase = wsend \\
  display = dup \\
  clientLock' = free \\
  displayed' = displayed \oplus (snapScene \cross \{ snapVal \}) \\
  batch' = batch \setminus snapScene \\
  drained' = drained \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  wphase' = wproc \\
  store' = store \\
  dirty' = dirty \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

On a send timeout (\texttt{BlockingIOError}) the display is alive but
wedged.  The send gives up within its time limit and releases the client
lock, and the worker turns to the reap-respawn path.  The batch is kept:
the respawn re-marks every live scene, which re-covers the scene in flight.

\begin{schema}{SendStuck}
  \Delta Repl
\where
  wphase = wsend \\
  display = dstuck \\
  clientLock' = free \\
  wphase' = wreap \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

On a dead peer (\texttt{OSError}) there is nothing to kill; the worker
closes the dead descriptor and reconnects.

\begin{schema}{SendDead}
  \Delta Repl
\where
  wphase = wsend \\
  display = ddead \\
  clientLock' = free \\
  wphase' = wreconn \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

% ============================================================
\section{Reap, Respawn, Reconnect}
% ============================================================

A stuck display still answers new connections, so it must be killed before
a fresh one is started.  \texttt{Reap} terminates the wedged process;
\texttt{Ensure} starts a fresh, empty one; \texttt{Remark} re-marks the work
so the display is repainted from the store's current state, re-marks the
clear if the failed batch carried one, and re-marks the menu flag if the
failed batch carried one.  \texttt{Ensure} is enabled only from
\texttt{ddead}: a fresh display is never spawned while one is alive, which is
I5.

The re-mark set is \texttt{drained} $\cup$ the live scenes: every scene the
failed batch drained, plus every scene the store still holds live.  The two
unions cover complementary gaps.  The live scenes must be re-marked because a
fresh or reconnected display shows none of them until they are pushed again.
The \emph{drained} scenes must be re-marked because a batch can drain a scene
that is \emph{not} live --- an emptied or departed scene whose only remaining
job is to be blanked; that blank can be lost to the failed send, and it is
absent from the live set, so re-marking only the live scenes would drop it.
Unioning \texttt{drained} keeps the emptied scene's blank in the queue until
it lands.  \texttt{SendRecovery.\_remark} does exactly this: it re-marks
\texttt{batch.scenes} together with \texttt{live\_scene\_ids()}.

The clear re-mark is the third part of the heal.  A cleared scene is empty in
the store, so it is absent from the live set, and \texttt{clear} marks no
scene dirty, so it is absent from \texttt{drained} as well; without re-marking
the clear, a reconnect to the same surviving display would leave the old,
pre-clear scene on screen forever.  Re-marking the clear re-arms the live
flag, so the next cycle blanks the display again before repainting;
blank-then-repaint is idempotent.  The re-mark is expressed against
\texttt{batchCleared} --- the clear this failed batch carried --- and merges
into whatever the live flag already holds, so a fresh clear that arrived
mid-cycle is never lost.

The menu re-mark is the fourth part, and it is the guarantor of I6.  A menu
send that failed left \texttt{menuShown = clear}, and the drained menu flag is
in \texttt{batchMenus}; re-marking it against \texttt{batchMenus} re-arms the
live \texttt{menusDirty} so the next cycle re-reads the registry fresh and
re-delivers.  \texttt{SendRecovery.\_requeue} does exactly this: it calls
\texttt{mark\_menus} when the batch's \texttt{menus\_dirty} was set.  Dropping
this re-mark is the fidelity variant of Section~\ref{sec:fidelity}, under
which a menu change whose send failed is neither delivered nor re-marked ---
an I6 violation.

\begin{schema}{Reap}
  \Delta Repl
\where
  wphase = wreap \\
  display' = ddead \\
  wphase' = wensure \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  shutting' = shutting
\end{schema}

\begin{schema}{Ensure}
  \Delta Repl
\where
  wphase = wensure \\
  display = ddead \\
  display' = dup \\
  displayed' = (\lambda s : SCENE @ vempty) \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  wphase' = wremark \\
  store' = store \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  shutting' = shutting
\end{schema}

\begin{schema}{Remark}
  \Delta Repl
\where
  wphase = wremark \\
  dirty' = (dirty \cup drained) \cup \{ s : SCENE | store~s = vlive \} \\
  (batchCleared = set \implies cleared' = set) \\
  (batchCleared = clear \implies cleared' = cleared) \\
  (batchMenus = set \implies menusDirty' = set) \\
  (batchMenus = clear \implies menusDirty' = menusDirty) \\
  batch' = \emptyset \\
  drained' = \emptyset \\
  wphase' = wproc \\
  store' = store \\
  displayed' = displayed \\
  batchCleared' = batchCleared \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\texttt{Reconn} handles the dead-peer path, and it is modelled adversarially.
A plain \texttt{OSError} does not mean the display \emph{process} died; it
means \emph{this connection} did.  The display is a long-lived process that
keeps its last-received scene copy and renders it every frame, so a reconnect
can land on the same still-running display, showing exactly what it showed
before the drop.  \texttt{Reconn} therefore leaves \texttt{displayed}
unchanged --- the worst case, stale content and all --- rather than pretending
a fresh blank display.  There is no process to kill, so it reconnects and
re-marks in one step: \texttt{drained} $\cup$ the live scenes, plus the clear
and the menu flag the batch carried.  I4 holds \emph{because of} that re-mark,
not because of any reset: every scene the failed cycle touched, and every live
scene, is re-queued, so the stale display cannot stay wrong once the queue
drains.

Modelling the reconnect this way is what makes the \texttt{drained} half of
the re-mark load-bearing.  Under a reset-to-empty reconnect a fresh display
would already blank every emptied scene, and re-marking \texttt{drained} would
change nothing an observer could see.  Against a stale-content display it is
the only thing that re-blanks an emptied scene whose blank was lost to the
failed send; Section~\ref{sec:fidelity} removes it and reproduces the linger.

\begin{schema}{Reconn}
  \Delta Repl
\where
  wphase = wreconn \\
  display' = dup \\
  displayed' = displayed \\
  dirty' = (dirty \cup drained) \cup \{ s : SCENE | store~s = vlive \} \\
  (batchCleared = set \implies cleared' = set) \\
  (batchCleared = clear \implies cleared' = cleared) \\
  (batchMenus = set \implies menusDirty' = set) \\
  (batchMenus = clear \implies menusDirty' = menusDirty) \\
  batch' = \emptyset \\
  drained' = \emptyset \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  wphase' = wproc \\
  store' = store \\
  batchCleared' = batchCleared \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  shutting' = shutting
\end{schema}

\texttt{CycleFail} is the generic failure outcome: any error inside the
cycle other than the two socket outcomes --- a display that will not spawn,
a reconnect that is refused, or any exception the two socket handlers do not
catch.  It can strike while the worker is sending a scene (\texttt{wsend}) or
the menu (\texttt{wmsend}) or mid-heal (\texttt{wreap}, \texttt{wensure},
\texttt{wreconn}).  The drained work must not be lost, so the worker re-marks
it --- \texttt{drained}, plus the clear and the menu flag the batch carried
--- releases whatever locks it held, and returns to \texttt{widle} to back off
and retry.  Restoring \texttt{drained} rather than the unsent remainder
\texttt{batch} is what saves an already-sent scene: if the batch carried a
clear, the re-marked clear will blank the whole display next cycle, so every
scene the batch drained --- including those already sent --- must be
repainted, or a painted scene would be blanked and never redrawn.

Unlike \texttt{Remark} and \texttt{Reconn}, \texttt{CycleFail} does \emph{not}
union the live scenes: it re-marks \texttt{drained} alone.  This matches
\texttt{SendRecovery.restore}, which re-queues \texttt{batch.scenes} without
\texttt{live\_scene\_ids}, and it is sound because \texttt{CycleFail} does not
replace the display.  A live scene the failed cycle did not touch is still
shown correctly on the same unblanked display, so it needs no repaint; only
the interrupted cycle's drained work does.  The live union is needed only when
the display is reset (\texttt{Remark}) or reconnected onto stale content
(\texttt{Reconn}), where a live scene may not be shown.

\begin{schema}{CycleFail}
  \Delta Repl
\where
  wphase \in \{ wsend, wmsend, wreap, wensure, wreconn \} \\
  dirty' = dirty \cup drained \\
  (batchCleared = set \implies cleared' = set) \\
  (batchCleared = clear \implies cleared' = cleared) \\
  (batchMenus = set \implies menusDirty' = set) \\
  (batchMenus = clear \implies menusDirty' = menusDirty) \\
  batch' = \emptyset \\
  drained' = \emptyset \\
  batchCleared' = clear \\
  batchMenus' = clear \\
  menuShown' = menuShown \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  readLock' = free \\
  clientLock' = free \\
  wphase' = widle \\
  store' = store \\
  displayed' = displayed \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  display' = display \\
  shutting' = shutting
\end{schema}

% ============================================================
\section{Ending a Cycle, Clear, Shutdown}
% ============================================================

A cleared cycle blanks the display \emph{before} it repaints the batch: the
\texttt{wblank} phase runs \texttt{ApplyClear} first, and the scenes are
reached only at \texttt{wproc}, so no scene is painted until the clear has
been applied.  The worker then repaints each batch scene from the store's
current value, so a \texttt{show} that coalesced after the \texttt{clear}
survives.

This is a deliberate departure from the original design text, which pushed
the clear \emph{after} the batch (\texttt{for scene in batch: push(...); if
was\_cleared: push\_clear()}).  Model-checking I4 against that order found a
lost update: a \texttt{clear} followed by a \texttt{show} within the
worker's 16\,ms coalescing window painted the new scene and then blanked it
(Section~\ref{sec:verify}).  The clear-first order is convergent for both
\texttt{show}-then-\texttt{clear} and \texttt{clear}-then-\texttt{show}, and
the shipped \texttt{\_attempt} implements it: \texttt{clear\_async} first,
then the menu push, then the per-scene repaint.  The clear's own send is
modelled as an immediate blank; its send has the same time limit and the same
failure outcomes as a scene send, which the send operations already cover.

When the batch is empty --- every scene sent or skipped --- the worker
drops the batch's carried clear and menu flag and goes idle.

\begin{schema}{ProcDone}
  \Delta Repl
\where
  wphase = wproc \\
  batch = \emptyset \\
  batchCleared' = clear \\
  batchMenus' = clear \\
  menuShown' = menuShown \\
  wphase' = widle \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = \emptyset \\
  cleared' = cleared \\
  menusDirty' = menusDirty \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

The display fails independently of the worker: an environment step may
wedge it or kill it.  These are the events the worker's next send detects.

\begin{schema}{DisplayWedge}
  \Delta Repl
\where
  display = dup \\
  display' = dstuck \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

\begin{schema}{DisplayCrash}
  \Delta Repl
\where
  display = dup \\
  display' = ddead \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  shutting' = shutting \\
  wphase' = wphase
\end{schema}

At clean shutdown the worker makes one last bounded flush of whatever is
still dirty and then stops.  The flush is best-effort: losing the last
frame is acceptable because the process is exiting.  \texttt{Restart}
re-enters a fresh idle state, modelling luxd starting again, and keeps the
model live so the deadlock check reports only genuine lock deadlocks, never
this deliberate terminal stop.

\begin{schema}{ShutdownReq}
  \Delta Repl
\where
  shutting = clear \\
  shutting' = set \\
  store' = store \\
  displayed' = displayed \\
  dirty' = dirty \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  wphase' = wphase
\end{schema}

\begin{schema}{Flush}
  \Delta Repl
\where
  wphase = widle \\
  shutting = set \\
  dirty' = \emptyset \\
  wphase' = wstopped \\
  store' = store \\
  displayed' = displayed \\
  batch' = batch \\
  drained' = drained \\
  cleared' = cleared \\
  batchCleared' = batchCleared \\
  menusDirty' = menusDirty \\
  batchMenus' = batchMenus \\
  menuShown' = menuShown \\
  snapScene' = snapScene \\
  snapVal' = snapVal \\
  storeLockMut' = storeLockMut \\
  mutScene' = mutScene \\
  readLock' = readLock \\
  clientLock' = clientLock \\
  display' = display \\
  shutting' = shutting
\end{schema}

\begin{schema}{Restart}
  \Delta Repl
\where
  wphase = wstopped \\
  store' = (\lambda s : SCENE @ vempty) \\
  displayed' = (\lambda s : SCENE @ vempty) \\
  dirty' = \emptyset \\
  batch' = \emptyset \\
  drained' = \emptyset \\
  cleared' = clear \\
  batchCleared' = clear \\
  menusDirty' = clear \\
  batchMenus' = clear \\
  menuShown' = set \\
  snapScene' = \emptyset \\
  snapVal' = vempty \\
  storeLockMut' = \emptyset \\
  mutScene' = \emptyset \\
  readLock' = free \\
  clientLock' = free \\
  display' = dup \\
  shutting' = clear \\
  wphase' = widle
\end{schema}

% ============================================================
\section{Invariants}
\label{sec:inv}
% ============================================================

Six properties must hold across all reachable states.  None is placed in
the \texttt{Repl} schema: a predicate placed there would be enforced as a
guard on every successor, silently disabling any operation that would break
it and thereby masking the very defect this model exists to catch.

\paragraph{Invariant 1 --- single writer.}
Only the worker ever holds the client send lock, and it holds it only while
sending --- a scene at \texttt{wsend} or the menu at \texttt{wmsend}:
\[
  clientLock = held \implies wphase \in \{ wsend, wmsend \}.
\]
No mutator operation mentions \texttt{clientLock} in its effect, so no
mutator can ever hold it.  The design removes the two second-writer paths
the current code has --- the inline resend in the click dispatch and the
beads app's direct render --- so that the worker's \texttt{wsend} and
\texttt{wmsend} are the only places a send happens.

\paragraph{Invariant 2 --- agent-path liveness.}
No mutator operation is guarded by \texttt{clientLock}, \texttt{display}, or
\texttt{wphase}: \texttt{ReplaceBegin}, \texttt{ReplaceCommit},
\texttt{EmptyScene}, \texttt{ClearStore}, and \texttt{MarkMenus} depend only on
the store write lock being free (and \texttt{MarkMenus} only on
\texttt{menusDirty}).  So no state of the display or of the worker's send can
disable a mutator.  A mutator's begin does also require \texttt{readLock =
free}, the ordinary reader-writer exclusion of the store lock; but the worker
holds that read lock only across \texttt{Snap}'s in-memory copy and releases
it in \texttt{SendBegin} before any send, so a mutator waits at most for one
snapshot copy and never for display I/O.  The witness
(Section~\ref{sec:verify}) is a reachable state in which the worker is stuck
sending to a wedged display while a mutator holds the write lock and is free
to commit.

\paragraph{Invariant 3 --- no lock cycle.}
No thread holds the store lock and the client send lock at once:
\[
  \lnot\,(readLock = held \land clientLock = held).
\]
The worker takes the read lock in \texttt{Snap}, releases it in
\texttt{SendBegin} \emph{before} taking the client lock, and no mutator
ever takes the client lock.  The menu send takes the client lock at
\texttt{wmsend} while holding no read lock, so it cannot form a cycle
either.  The two locks are therefore always acquired in one order and never
held together, so no cyclic wait can form.

\paragraph{Invariant 4 --- no lost update.}
Whenever the system is at rest with the display up, the display shows the
store's latest value for every scene.  ``At rest'' means the worker is
idle, nothing is dirty, cleared, or menu-marked, no mutator holds the write
lock, and no shutdown is in progress:
\[
  Quiescent \implies (\forall s : SCENE @ displayed~s = store~s),
\]
where
\begin{eqnarray*}
  Quiescent & \defs & wphase = widle \land dirty = \emptyset \land
    cleared = clear \\
  & & {} \land menusDirty = clear \land storeLockMut = \emptyset \land
    shutting = clear \land display = dup.
\end{eqnarray*}
Every recovery re-marks at least \texttt{drained} plus the carried clear, so
the system cannot come to rest until that work has been pushed, which restores
\texttt{displayed = store}.  The two display-replacing recoveries also union
the live scenes: \texttt{Remark} follows a respawn onto a fresh, empty
display, and \texttt{Reconn} follows a reconnect onto a stale-content display,
and in both a live scene may not be shown, so re-marking it is what closes
that gap.  \texttt{CycleFail} does not replace the display, so a live scene it
did not touch is still shown correctly and needs no repaint; it re-marks
\texttt{drained} alone.  In every case the re-mark covers the two gaps that
can actually open --- a live scene a replaced display is not showing, and an
emptied scene whose blank was lost --- so no scene can stay wrong at rest.  An emptied scene reaches rest only once its blank has
been pushed (\texttt{SendOk} with \texttt{snapVal = vempty}), or, under a
clear, once the whole-display blank has set it empty; either way
\texttt{displayed~s = store~s = vempty}.  \texttt{batchCleared} does not
appear in \texttt{Quiescent} because it is \texttt{clear} at every
\texttt{widle} state: every path into \texttt{widle} --- \texttt{ProcDone},
\texttt{CycleFail}, \texttt{Init}, \texttt{Restart} --- resets it.

\paragraph{Invariant 5 --- one display.}
A fresh display is spawned only after the old one is killed:
\texttt{Ensure} is enabled only from \texttt{display = ddead}, and
\texttt{Reap} is the only step that reaches \texttt{ddead} on the recovery
path.  So the worker never holds two live displays.

\paragraph{Invariant 6 --- menu-delivery durability.}
Whenever the system is at rest with the display up, no marked menu change is
left undelivered:
\[
  Quiescent \implies menuShown = set.
\]
\texttt{menuShown} goes \texttt{clear} only at \texttt{MarkMenus}, which also
sets \texttt{menusDirty}; it goes \texttt{set} only at \texttt{MenuSendOk}, a
successful push.  A \texttt{Quiescent} state requires \texttt{menusDirty =
clear}, and the only way \texttt{menusDirty} returns to \texttt{clear} after a
mark is \texttt{Drain}, which carries the change into \texttt{batchMenus}.
That drained change is then either delivered --- \texttt{SendMenusBegin} then
\texttt{MenuSendOk}, which sets \texttt{menuShown = set} --- or its send fails,
and every failure path re-marks \texttt{menusDirty} against
\texttt{batchMenus} (\texttt{Remark}, \texttt{Reconn}, \texttt{CycleFail}),
which re-arms the live flag and so denies \texttt{Quiescent} until a later
cycle re-delivers.  Hence a rest state with the display up cannot be reached
with \texttt{menuShown = clear}: the marked change was pushed, or it is still
marked and the system is not at rest.  This is the delivery obligation of the
replicator alone; that a freshly respawned display also re-obtains its
registered items over the client handshake is a separate mechanism the model
does not rely on.

% ============================================================
\section{Verification}
\label{sec:verify}
% ============================================================

Type-checking is a \texttt{fuzz} pass:
\begin{verbatim}
  fuzz docs/hub_replicator.tex
\end{verbatim}

I3, I4, I5, and I6 are state properties, checked by asking ProB whether their
negation is \emph{reachable}.  A goal reported \emph{not found} means the
invariant holds on every reachable state:
\begin{verbatim}
  # Invariant 3 (must report: goal NOT found)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "readLock = held & clientLock = held" \
    -p DEFAULT_SETSIZE 2

  # Invariant 4 (must report: goal NOT found)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = widle & dirty = {} & cleared = clear &
           menusDirty = clear & storeLockMut = {} & shutting = clear &
           display = dup &
           card({s | s : SCENE & displayed(s) /= store(s)}) > 0" \
    -p DEFAULT_SETSIZE 2

  # Invariant 5 (must report: goal NOT found)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wensure & display /= ddead" \
    -p DEFAULT_SETSIZE 2

  # Invariant 6 (must report: goal NOT found)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = widle & dirty = {} & cleared = clear &
           menusDirty = clear & storeLockMut = {} & shutting = clear &
           display = dup & menuShown = clear" \
    -p DEFAULT_SETSIZE 2
\end{verbatim}

I1 is corroborated by a reachability goal that must be \emph{not found} ---
no state holds the client lock outside a scene or menu send:
\begin{verbatim}
  # Invariant 1 (must report: goal NOT found)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "clientLock = held & wphase /= wsend & wphase /= wmsend" \
    -p DEFAULT_SETSIZE 2
\end{verbatim}

I2 is a liveness witness that must be \emph{found}: a reachable state in
which the worker is sending to a wedged display while a mutator holds the
write lock and can still commit --- the agent path is not blocked by
display I/O:
\begin{verbatim}
  # Invariant 2 (must report: goal FOUND)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "display = dstuck & clientLock = held & storeLockMut /= {}" \
    -p DEFAULT_SETSIZE 2
\end{verbatim}

Deadlock freedom is the full-state-space check.  \texttt{Restart} keeps the
model live, so any deadlock reported would be a genuine lock deadlock:
\begin{verbatim}
  # Deadlock freedom (must report: no deadlock / no counterexample)
  probcli docs/hub_replicator.tex -model_check -p DEFAULT_SETSIZE 2
\end{verbatim}

The recovery-and-blanking behaviours and the menu leg must be reachable, or
their invariants would hold only because the machinery is dead.  Each of these
witnesses must be \emph{found}:
\begin{verbatim}
  # Generic cycle failure is reachable and restores its batch
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wsend & batch /= {}" -p DEFAULT_SETSIZE 2

  # A batch that carried a clear reaches recovery (clear re-mark is live)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wremark & batchCleared = set" -p DEFAULT_SETSIZE 2

  # The empty-scene-under-clear skip is reachable
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wsnap & snapVal = vempty & batchCleared = set" \
    -p DEFAULT_SETSIZE 2

  # An emptied scene is blanked without a clear (behaviour 3, happy path)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wsend & snapVal = vempty & batchCleared = clear" \
    -p DEFAULT_SETSIZE 2

  # The menu send is reachable (the menu leg is live)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wmsend & batchMenus = set" -p DEFAULT_SETSIZE 2

  # A menu-carrying batch reaches recovery (the menu re-mark is live)
  probcli docs/hub_replicator.tex -model_check -nodead \
    -goal "wphase = wremark & batchMenus = set" -p DEFAULT_SETSIZE 2
\end{verbatim}

\paragraph{Finding --- clear ordering.}
Model-checking I4 against the original design document's stated order ---
push the batch scenes, then \texttt{if was\_cleared: push\_clear()} ---
returns \emph{goal found}.  Replace the \texttt{wblank}-phase blank (which
happens first) with a \texttt{PushClear} that blanks at the end of the
cycle, and this trace is reachable:
\begin{verbatim}
  ClearStore              ; store emptied, cleared set
  ReplaceBegin/Commit s1  ; a show() lands after the clear, within
                          ;   the 16ms coalescing window; store s1 = vlive
  Drain                   ; batch = {s1}, batchCleared = set
  Snap; SendBegin; SendOk ; display now shows s1 = vlive
  PushClear               ; display blanked to empty
  -- rest: displayed s1 = vempty, store s1 = vlive  (lost update)
\end{verbatim}
A \texttt{clear} immediately followed by a \texttt{show} is an ordinary
sequence, so this is a reachable defect, not a corner case.  The clear-first
order --- the \texttt{wblank} phase preceding \texttt{wproc} --- closes it:
with clear-first the goal returns \emph{not found}, and the shipped
\texttt{\_attempt} sends \texttt{clear\_async} before the per-scene repaint.

\paragraph{Note --- the two reconnect models.}
The reap path and the dead-peer path differ in one honest way, and the model
keeps them apart.  \texttt{Reap}/\texttt{Ensure} kill the wedged process and
spawn a new one, which genuinely starts blank, so the model resets
\texttt{displayed} to empty.  \texttt{Reconn} does not spawn anything: an
\texttt{OSError} kills the connection, not necessarily the display process,
and that process keeps its last-received copy and renders it every frame.
The model therefore leaves \texttt{displayed} unchanged on \texttt{Reconn} ---
the adversarial, faithful case of a reconnect onto a still-running display
that still shows the pre-failure scene.

That choice is what makes the \texttt{drained} half of the re-mark
load-bearing.  Had \texttt{Reconn} reset to empty, a fresh display would
already blank every emptied scene, and unioning \texttt{drained} into the
re-mark would change nothing an observer could see --- the model would verify
clean while being too abstract to guard the fix.  Against a stale-content
reconnect it is exactly the union that saves an emptied scene: its blank,
lost to the failed send, is absent from the live set but present in
\texttt{drained}, so re-marking \texttt{drained} $\cup$ live re-queues it.
Section~\ref{sec:fidelity} removes that union and reproduces the linger the
fix closes.  This is the gap flagged in the previous revision of this report,
now closed in both the code (\texttt{SendRecovery.\_remark} unions
\texttt{batch.scenes}) and the model.

% ============================================================
\section{Fidelity: reproducing the known defects}
\label{sec:fidelity}
% ============================================================

A model that cannot reproduce the known defects proves nothing.  This
specification guards three, and removing any one fix makes the corresponding
defect reachable.

\paragraph{The store race.}
The first buggy variant removes the store's reader-writer exclusion from the
worker's snapshot.  Concretely, delete the guard \texttt{storeLockMut =
$\emptyset$} from \texttt{Snap}, so the worker can copy a scene's value while
a mutator is inside \texttt{replace\_scene} and the scene is \texttt{vtorn}.
The torn read then becomes reachable:

\begin{enumerate}
  \item Mutator $m$ runs \texttt{ReplaceBegin} on scene $s$: it takes the
    write lock and sets \texttt{store~$s$ = vtorn}, with
    \texttt{storeLockMut = \{m\}}.
  \item Meanwhile the worker has a batch containing $s$ and is at
    \texttt{wproc}.  Without the exclusion guard, \texttt{Snap} fires: it
    copies \texttt{snapVal = store~$s$ = vtorn}.
  \item The worker is now about to send a half-updated scene: a state with
    \texttt{snapVal = vtorn}.
\end{enumerate}

The property that separates the two designs is therefore
\begin{verbatim}
  # must be NOT found for the correct spec, FOUND for the buggy variant
  probcli <spec>.tex -model_check -nodead \
    -goal "snapVal = vtorn" -p DEFAULT_SETSIZE 2
\end{verbatim}
Under the intact exclusion the worker's \texttt{Snap} cannot fire while any
mutator holds the write lock, so \texttt{snapVal} is never \texttt{vtorn};
the reachability search returns \emph{goal not found}.  With the guard
removed it returns \emph{goal found} --- the trace above.  That difference
is the evidence that the model detects the race the store lock was added to
prevent.

\paragraph{The emptied-scene linger.}
The second buggy variant removes the \texttt{drained} half of the recovery
re-mark.  Concretely, in \texttt{Reconn} replace
\texttt{dirty$'$ = (dirty $\cup$ drained) $\cup$ \{s | store~s = vlive\}}
with the live-only \texttt{dirty$'$ = dirty $\cup$ \{s | store~s = vlive\}}
--- the pre-fix recovery that re-marked only the live scenes.  With
\texttt{Reconn} modelled against a stale-content display, an emptied scene's
lost blank is then never re-queued:

\begin{enumerate}
  \item \texttt{show} paints scene $s$: \texttt{store~$s$ = vlive},
    \texttt{displayed~$s$ = vlive}.
  \item \texttt{EmptyScene}~$s$: \texttt{store~$s$ = vempty}, $s$ marked
    dirty.  Its blank is now the worker's job.
  \item The worker drains $s$ and sends its blank; the send raises
    \texttt{OSError}.  \texttt{SendDead} $\to$ \texttt{wreconn}.
  \item \texttt{Reconn} leaves \texttt{displayed~$s$ = vlive} (stale
    display) and, live-only, re-marks nothing for $s$ --- $s$ is empty in the
    store, so it is not live.  $s$ is dropped.
  \item At rest: \texttt{displayed~$s$ = vlive}, \texttt{store~$s$ = vempty}
    --- a \texttt{Quiescent} lost update.
\end{enumerate}

The property that separates the two designs is I4 itself:
\begin{verbatim}
  # must be NOT found for the correct spec, FOUND for the live-only variant
  probcli <spec>.tex -model_check -nodead \
    -goal "wphase = widle & dirty = {} & cleared = clear &
           menusDirty = clear & storeLockMut = {} & shutting = clear &
           display = dup &
           card({s | s : SCENE & displayed(s) /= store(s)}) > 0" \
    -p DEFAULT_SETSIZE 2
\end{verbatim}
With the \texttt{drained} union intact the emptied scene is re-queued and
reblanked, so the goal returns \emph{not found}.  With it removed the goal
returns \emph{found} --- the trace above.  That difference is the evidence
that the model detects the defect reviewers found in the live-only recovery,
which \texttt{SendRecovery.\_remark}'s union of \texttt{batch.scenes} closes.

\paragraph{The dropped menu change.}
The third buggy variant removes the menu re-mark from the recovery.
Concretely, drop the two lines
\texttt{(batchMenus = set $\implies$ menusDirty$'$ = set)} and
\texttt{(batchMenus = clear $\implies$ menusDirty$'$ = menusDirty)} from
\texttt{Remark}, \texttt{Reconn}, and \texttt{CycleFail}, replacing them with
an unconditional \texttt{menusDirty$'$ = menusDirty} --- the recovery that
re-queues scenes and the clear but forgets the menu.  A menu change whose send
failed is then neither delivered nor re-marked:

\begin{enumerate}
  \item \texttt{MarkMenus}: \texttt{menusDirty = set}, \texttt{menuShown =
    clear} --- a menu change is owed.
  \item \texttt{Drain}: \texttt{batchMenus = set}, \texttt{menusDirty = clear}.
  \item \texttt{SkipBlank}; \texttt{SendMenusBegin}: the worker takes the
    client lock to push the menu, reaching \texttt{wmsend}.
  \item The display is wedged (\texttt{DisplayWedge} beforehand), so
    \texttt{MenuSendStuck} fires: \texttt{wreap}, \texttt{menuShown} still
    \texttt{clear}.
  \item \texttt{Reap}; \texttt{Ensure}; \texttt{Remark} --- but, live-only, it
    does not re-mark \texttt{menusDirty}.  \texttt{batchMenus} is consumed.
  \item No scenes remain, so \texttt{ProcDone} returns the worker to
    \texttt{widle}.
  \item At rest: \texttt{menusDirty = clear}, \texttt{menuShown = clear} ---
    a \texttt{Quiescent} state with a menu change that was marked, never
    delivered, and no longer marked.  I6 is violated.
\end{enumerate}

The property that separates the two designs is I6 itself:
\begin{verbatim}
  # must be NOT found for the correct spec, FOUND for the no-remark variant
  probcli <spec>.tex -model_check -nodead \
    -goal "wphase = widle & dirty = {} & cleared = clear &
           menusDirty = clear & storeLockMut = {} & shutting = clear &
           display = dup & menuShown = clear" \
    -p DEFAULT_SETSIZE 2
\end{verbatim}
With the menu re-mark intact, a failed menu send re-arms \texttt{menusDirty},
so the system cannot rest until a later cycle re-reads the registry and
re-delivers; the goal returns \emph{not found}.  With it removed the goal
returns \emph{found} --- the trace above.  That difference is the evidence
that the model detects a dropped menu change, which
\texttt{SendRecovery.\_requeue}'s call to \texttt{mark\_menus} closes.

\end{document}
